### 2024-10-31 A3-Pt1.Rmd
Add files via upload

---
title: "Exploring Disparate Data: Part 1"
author: "Put your group number from MyLS here: "
date: "Due November 12th"
---

```{r active="", eval=FALSE}
# BEGIN ASSIGNMENT 
```

```{r}
knitr::opts_chunk$set(error = TRUE)
# DO NOT PUT install.packages() IN AN RMD!!!
# While you're here, make sure you group number is filled in above.
    # (And make sure it ends with a quotation mark).
library(tidyverse)
theme_set(theme_bw())
library(arrow)
library(openxlsx)
```

List all group members here:

- 
- 
- 
- 
-

This file will be submitted as an Rmd file to GradeScope. You will need to create a group on GradeScope with the correct members. *If the person who submits this does not list you, you will not get a grade on this assignment!!!* Make sure you trust the person who is submitting it!

## Overview

This is **Part 1** of the group assignment, in which you'll demonstrate your ability to clean the data sets.

In this part, you will prepare the following data sets for analysis in Part 2. Note that you'll only need to use three of them, so you can focus on those three for cleaning and get started on Part 2.

Once you've completed the relevant sections, you will have the following files in your working directory:

1. `cyclones_data.parquet` contains data for hurricane strength in both the Atlantic and North Pacific basins.
2. `ice_extent_yearly.parquet` contains the yearly ice extent for the Arctic and the Antarctic poles.
3. `climate_awareness.parquet` contains the proportion of people from each country who answered "no", "a little", ... to a question asking about their awareness of the actual definitions of climate change.
4. `covid_2020.parquet` contains the total reported cases of COVID-19 in the year 2020 for all countries.
    - It also, quite helpfully, includes the continent on which each country sits, which can be joined with other data frames (with some care).
5. `happiness.parquet` contains information about the happiness of countries in the world, as measured by the `life_ladder` question (among other survey responses).

## NOAA data for Atlantic and Pacific Basin

### Load in the Data

In assignment 2, the NOAA data for the Atlantic was cleaned and tidied. That data will be useful for this assignment, so re-do the steps here. In this version, all of the steps can be done in a single pipeline. We'll also be doing the North Pacific (NP) basin.

```{r}
cyclone_data_address <- "https://www.nhc.noaa.gov/data/hurdat/"
at_cyclone_filename <- "hurdat2-1851-2022-050423.txt"
np_cyclone_filename <- "hurdat2-nepac-1949-2022-050423.txt"

new_columns <- c("status", "latitude", "longitude", "max_wind",
    "min_pressure", "NE_extend_34", "SE_extend_34", "SW_extend_34",
    "NW_extend_34", "NE_extend_50", "SE_extend_50", "SW_extend_50",
    "NW_extend_50", "NE_extend_64", "SE_extend_64", "SW_extend_64",
    "NW_extend_64", "r_max_wind"
)
```

```{r at_cyclone, error=TRUE}

# Same steps as in A2, but you can put them all in the same pipeline!
at_cyclone <- str_c(cyclone_data_address, at_cyclone_filename, sep = "") |>
    read_csv(
        col_names = c(as.character(1:4)),
        progress = FALSE,
        show_col_types = FALSE
    ) |>
    separate_wider_delim(
        cols = `4`,
        # Set the delim and the names
        # YOUR CODE HERE
    ) |>
    mutate(
        across(everything(), str_trim),
        # make "-999" NAs, make "-99" NAs
        # Create columns BasinNumberYear, Name, and Entries
        across(everything(), ~na_if(., "-999")),
        across(everything(), ~na_if(., "-99")),
        BasinNumberYear = ifelse(is.na(status), `1`, NA),
        Name = ifelse(is.na(status), `2`, NA),
        Entries = ifelse(is.na(status), `3`, NA)
    ) |>
    relocate(BasinNumberYear, Name, Entries) |>
    fill(BasinNumberYear, Name, Entries) |>
    filter(!is.na(status))  |>
    select(-Entries) |>
    separate_wider_position(
        BasinNumberYear,
        # Specify the widths
        # YOUR CODE HERE
    ) |>
    separate_wider_position(
        `1`,
        # Specify the widths
        # YOUR CODE HERE
    ) |>
    separate_wider_position(
        `2`,
        # Specify the widths
        # YOUR CODE HERE
    ) |>
    rename(
        Identifier = `3`
    ) |>
    mutate(
        across(
            c(NameYear, ObservYear, Month, Day, Hour,
                Minute, Number),
            as.integer
        )
    ) |>
    mutate(across(max_wind:r_max_wind, as.numeric))

print(at_cyclone)
```

```{r}
. = ottr::check("tests/at_cyclone.R")
```

```{r}
np_cyclone <- str_c(cyclone_data_address, np_cyclone_filename, sep = "") |>
    # ALL of the steps all over again
    # YOUR CODE HERE

print(np_cyclone)
```

```{r}
. = ottr::check("tests/np_cyclone.R")
```

### Combine into One Data Frame

Now, bind the two data frames together (row-wise). Since the `Basin` column already contains the information about which basin the cyclone was in, there's no need to specify an ID. (Hint: don't overthink this one.)

Name the data frame `cyclones_data_update_0`.

```{r error=TRUE}
# YOUR CODE HERE

print(cyclones_data_update_0)
```

```{r}
. = ottr::check("tests/cyclones_data_update_1.R")
```

### Fix Latitude and Longitude

The latitude and longitude are not in a format that makes for nice plotting. They're encoded as "`28.0N`" for 28 degrees North and "28.0S" for 28 degrees South of the equator. By convention, "`28.0N`" should be `28` degrees (a positive number) and `28.8S` should be -28 degrees (a negative number). Similarly, `94W` should be -94 while 94E should be 94. Fix this.

Write a function that takes in a vector of latitude _or_ longitude in the format of these data, then outputs a positive version of the numeric part if there's an "N" or an "E", and a negative version of the numeric part if there's a "W" or "S".

There are many ways to solve this. My solution is one line and uses `parse_number()` and a single `if_else()` statement with a regex, but there are many possible ways to do this!

```{r convert_latlon, error=TRUE}
convert_latlon <- function(latlon) {
    # YOUR CODE HERE
}

test_data <- c("49W", "49.99W", "49E", "49.99E", "49N", "49.99S", "-0.0W")

# If this returns "TRUE", then your function will work on the real data.
all.equal(
    convert_latlon(test_data),
    c(-49, -49.99, 49, 49.99, 49, -49.99, 0)
)

cyclones_data_update_1 <- cyclones_data_update_0 |>
    mutate(
        lat = convert_latlon(latitude),
        lon = convert_latlon(longitude)
    )
```

```{r}
. = ottr::check("tests/convert_latlon.R")
```

### Extract Dates

We have a few more tasks. First, create a [datetime](https://fralfaro.github.io/r4ds/datetimes.html) object using the `make_datetime()` function so that R knows how to plot the times. Label the new column as "`date`".

To do this, you must choose whether to use `ObservYear` or `NameYear`.

```{r cyclones_data_update_2, error=TRUE}
cyclones_data_update_2 <- cyclones_data_update_1 |>
    mutate(
        # YOUR CODE HERE
    )

print(cyclones_data_update_2)
```

```{r}
. = ottr::check("tests/cyclones_data_update_2.R")
```

### Assign Storm Categories

Unlike in assignment 1, the category is not part of these data. However, we learned how storms are categorized! In particular, meteorologists use the [Saffir--Simpson hurricane wind scale (SSHWS)](https://en.wikipedia.org/wiki/Saffir%E2%80%93Simpson_scale) (note that our data are measured in knots on that scale).

Categorize the cyclones based on the scale above using a `case_when()` statement based on the `max_wind` column. The code is set up to create an [*ordered factor*](https://fralfaro.github.io/r4ds/factors.html) so that R knows that the levels are ordered as TD < TS < 1 < 2 < 3 < 4 < 5.
```{r cyclones_data}
cat_levels <- c("TD", "TS", "1", "2", "3", "4", "5")

cyclones_data <- cyclones_data_update_2 |>
    mutate(
        category = ordered(
            case_when(
                # YOUR CODE HERE
            ),
            levels = cat_levels
        )
    )

print(cyclones_data)
```

```{r}
. = ottr::check("tests/cyclones_update_2.R")
```

### Plots and Summary Statistics!

### Clean Up

Now that we have a suitable data frame, save it as a parquet file with the file name `"cyclones_data.parquet"`.
```{r}
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/write_cyclones.R")
```

## Sea Ice Data

The following data are used in the Course Notes to demonstrate reading in specific sheets. In this assignment, we care most about what's *in* the data! I've copied the code over from the notes; there are no questions on loading the data.

The sea ice extent is [defined](https://arctic.noaa.gov/report-card/report-card-2023/sea-ice-2023/) as the total area of the ocean that is covered in ice of at least 15% concentration.

```{r}
sea_ice_extent_xlsx <- "https://masie_web.apps.nsidc.org/pub//DATASETS/NOAA/G02135/seaice_analysis/Sea_Ice_Index_Daily_Extent_G02135_v3.0.xlsx"
```

```{r}
NH_daily <- sea_ice_extent_xlsx |>
    read.xlsx(
        sheet = "NH-Daily-Extent",
    ) |>
    select(X1, X2, `1978`:`2023`) |>
    rename(
        month = X1,
        day = X2
    ) |>
    fill(month) |>
    pivot_longer(
        cols = `1978`:`2023`,
        names_to = "year",
        values_to = "ice_extent",
        values_drop_na = TRUE,
    ) |>
    mutate(
        year = as.integer(year),
        month = ordered(
            month,
            levels = c("January", "February", "March", "April",
                "May", "June", "July", "August", "September",
                "October", "November", "December")),
        region = "Arctic",
    ) |>
    arrange(
        year, month, day
    )
```

```{r}
SH_daily <- sea_ice_extent_xlsx |>
    read.xlsx(
        sheet = "SH-Daily-Extent",
        skipEmptyCols = TRUE,
        fillMergedCells = TRUE,
        cols = 1:48
    ) |>
    rename(
        month = X1,
        day = X2
    ) |>
    pivot_longer(
        cols = `1978`:`2023`,
        names_to = "year",
        names_transform = list(year = as.integer),
        values_to = "ice_extent",
        values_drop_na = TRUE,
    ) |>
    mutate(
        month = ordered(
            month,
            levels = c("January", "February", "March", "April",
                "May", "June", "July", "August", "September",
                "October", "November", "December")
        ),
        region = "Antarctic",
    ) |>
    arrange(
        year, month, day
    )
```

```{r}
ice_extent_daily <- bind_rows(NH_daily, SH_daily) |>
    mutate(date = make_date(year, month, day)) |>
    arrange(region, date)

ice_extent_daily
```

## Stats and Plots

```{r}
ice_extent_daily |>
    ggplot() +
        aes(x = yday(date), y = ice_extent, colour = year, group = factor(year)) +
        geom_line() +
        facet_wrap(~region) +
        #coord_polar() +
        scale_colour_distiller(
            direction = 1, type = "seq", palette = 3
        )
```

In the chunk below, find the maximum and minimum sea ice extent for each year (by region). Pivot the data frame so that there's a column labelled "name" and a column labelled "value", as below.

| year  | region    | name | value   |
|-------|-----------|------|---------|
| 1978  | Antarctic | min  | 7.283   |
| 1978  | Antarctic | max  | 17.803  |
| 1978  | Arctic    | min  | 10.231  |
| 1978  | Arctic    | max  | 14.585  |
| 1979  | Antarctic | min  | 2.911   |
| 1979  | Antarctic | max  | 18.361  |
| ...   | ...       | ...  | ...     |

```{r}
ice_extent_yearly <- ice_extent_daily |>
    # YOUR CODE HERE

ice_extent_yearly
```

```{r}
. = ottr::check("tests/ice_extent_yearly.R")
```

Given your answer, the following code should produce a plot thats easier to interpret than the one before.

```{r plot_sea_ice_extent}
ggplot(ice_extent_yearly) +
    aes(x = year, y = value, colour = name) +
    geom_line() +
    facet_wrap(~ region) +
    labs(
        x = "Year", y = "Sea Ice Extent",
        colour = "Stat",
        title = "Min and Max Sea Ice Extent, by Year",
        subtitle = "Arctic is clearly decreasing, Antarctic is possibly becoming more variable."
    )
```

Finally, let's save the data to `ice_extent_yeary.parquet`.

```{r sea_ice_parquet, error=TRUE}
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/sea_ice_parquet.R")
```

# Opinions on climate data from Meta (from A2)

These data are the same as the data we worked with on A2. The following code will load in the awareness data; you are welcome to copy it for the happening data for the purpose of this assignment (you will not be graded on this, but it may benefit you in part 2).

```{r}
climate_opinion_address <- "https://data.humdata.org/dataset/dc9f2ca4-8b62-4747-89b1-db426ce617a0/resource/6041db5f-8190-47ff-a10b-9841325de841/download/climate_change_opinion_survey_2022_aggregated.xlsx"

climate_sheet_names <- climate_opinion_address |>
    loadWorkbook() |>
    names()

aware_sheet_name <- "climate_awareness"

climate_awareness <- climate_opinion_address |>
    read.xlsx(
        sheet = aware_sheet_name
    ) |>
    pivot_longer(
        cols = !contains(aware_sheet_name),
        names_to = "country",
        values_to = "score"
    ) |>
    mutate(
        climate_awareness = case_when(
            climate_awareness == "I have never heard of it" ~ "aware_no",
            climate_awareness == "I know a little about it" ~ "aware_alittle",
            climate_awareness == "I know a moderate amount about it" ~
                "aware_moderate",
            climate_awareness == "I know a lot about it" ~ "aware_alot",
            climate_awareness == "Refused" ~ "aware_refuse",
            climate_awareness == "(Unweighted Base)" ~ "aware_base"
        )
    ) |>
    rename(answer = climate_awareness) |>
    pivot_wider(
        names_from = answer,
        values_from = score
    )

write_parquet(climate_awareness, "climate_awareness.parquet")

climate_awareness
```

# COVID Data from Our World In Data

Perhaps theres some relationship between governments' response to the pandemic and governments' response to climate change? The following data will load in some information about COVID-19 in countries around the world.

```{r}
owid_address <- "https://covid.ourworldindata.org/data/owid-covid-data.csv"
covid_cases <- owid_address |>
    read_csv(
        col_types = cols(
            .default = col_double(),
            date = col_date(format = ""),
            iso_code = col_character(),
            location = col_character(),
            continent = col_character(),
            tests_units = col_character()
        ),
        progress = FALSE,
        show_col_types = FALSE
    )

covid_cases
```

The data are collected daily, which leads to a lot of data! This is probably more than we need to compare to the other variables.

In the chunk below, aggregate the data to be total cases in 2020, by country. Your solution should only have columns `country`, `continent` and `total_cases`. (*Hint*: The `year()` function will extract the year from a date column.)

```{r covid_2020, error=TRUE}
# YOUR CODE HERE

covid_2020
```

```{r}
. = ottr::check("tests/covid_2020.R")
```

Before we move on, we'll save `covid_2020.parquet` for use in Part 2.

```{r covid_2020_parquet, error=TRUE}
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/covid_2020_parquet.R")
```

As a very early exploration, let's compare climate awareness to the total number of covid cases in a country *per capita* in the year 2020.

In order to calculate per capita values, we need the population size. The easiest way that I can find is with the R package [`wbstats`](https://cran.r-project.org/web/packages/wbstats/vignettes/wbstats.html), which pulls updated statistics from the World Bank.

```{r}
# DO NOT PUT install.packages() IN AN RMD
# Run install.packages("wbstats") in your console, not in this code chunk
library(wbstats)
pop_data <- wb_data("SP.POP.TOTL", start_date = 2020, end_date = 2020)

```

A quick check to see which countries are in one data set but not the other:

```{r}
pop_countries <- pop_data |> pull(country)
aware_countries <- climate_awareness |> pull(country)

print("Countries in pop_data that aren't in climate_awareness:")
setdiff(pop_countries, aware_countries)
print("Countries in aware_countries that aren't in pop_data:")
setdiff(aware_countries, pop_countries)
```

Many of these countries are actually in both, but have a different name! Before we can join these two data sets together, we need the primary key (`country`) to match where possible.

- One pattern that I notice is that aware_countries has periods (`.`) instead of spaces. As a first step (to save us from manually entering 29 different country names twice), change all periods to spaces using `str_replace()`.
    - NOTE: The `.` is a special character in regex and needs to be [escaped](https://r4ds.hadley.nz/regexps#sec-regexp-escaping)!

```{r aware_fixed_country, error=TRUE}
aware_fixed_country <- climate_awareness |>
    mutate(
        # Replace *all* *literal* periods with a space in the country column
        # YOUR CODE HERE
    )

aware_fixed_country |> pull(country) |> unique()
```

```{r}
. = ottr::check("tests/aware_fixed_country.R")
```

Now let's run the check again:

```{r}
pop_countries <- pop_data |> pull(country)
aware_countries <- aware_fixed_country |> pull(country)

print("Countries in pop_data that aren't in climate_awareness:")
setdiff(pop_countries, aware_countries) |> sort()
print("Countries in aware_countries that aren't in pop_data:")
setdiff(aware_countries, pop_countries) |> sort()
```

That's slightly less that we need to deal with! Overall, the `pop_data` has more countries in it than `aware_fixed_country`, so we'll just do our best to change the countries in `aware_fixed_country` to match `pop_data`. Note that there a lot of [controversial politics](https://www.youtube.com/watch?v=3nB688xBYdY) involved in defining a "country" (esp. Taiwan, Hong Kong, West Bank and Gaza, etc.), and I'll avoid these as much as possible for now.

Use a `case_when()` statement to change the countries in `aware_fixed_country` where possible. Specifically, fix the following:

- "Bolivia (Plurinational State of)", "Congo (Democratic Republic of the)", "Egypt", "Hong Kong", "Lao People's Democratic Republic", "Slovakia", "North MAcedonia", "South Korea" (be careful with this one!), "Turkey", "Vietnam", "Yemen".
```{r aware_fixed_country_match}
aware_fixed_country_match <- aware_fixed_country |>
    mutate(
        # Change country using a case_when statement
        # YOUR CODE HERE
    )

aware_countries <- aware_fixed_country_match |> pull(country)
print("Countries in aware_countries that aren't in pop_data:")
setdiff(aware_countries, pop_countries) |> sort()
```

```{r}
. = ottr::check("tests/aware_fixed_country_match.R")
```

The remaining coutries are either broader geopolitical areas rather than countries, or Taiwan (which is apparently recognized as a country by Meta but not World Bank). For simplicity, we will simply move on without these countries/areas.

Now we can join! Use a [join](https://dplyr.tidyverse.org/reference/mutate-joins.html) that only keeps countries that are in both data sets. The final data frame should have the columns `country`, `total_population`, `aware_no`, `aware_alittle`, `aware_moderate`, `aware_alot`, and `aware_refuse`.

```{r aware_pop, error=TRUE}
# YOUR CODE HERE

aware_pop
```

```{r}
. = ottr::check("tests/aware_pop.R")
```

Finally, let's make sure the country names in `covid_2020` match those in `aware_pop`. In the end, all countries in `covid_2020` should be in `aware_pop` (but `aware_pop` will have countries that are not present in `covid_2020_match_country`, which is unavoidable).

```{r}
aware_countries <- aware_pop |> pull(country) |> unique()
covid_countries <- covid_2020 |> pull(country) |> unique()

setdiff(covid_countries, aware_countries)
setdiff(aware_countries, covid_countries)
```

```{r covid_2020_match_country, error=TRUE}
# YOUR CODE HERE

covid_2020_match_country
```

```{r}
. = ottr::check("tests/covid_2020_match_country.R")
```

Okay, now take a quick breather. Grab a tea, go for a walk, rock out to some music, whatever self care you want. That was a lot of finicky stuff, but now we've got a data set that we can use!

Join `aware_pop` and `covid_2020_match_country`, again only keep countries that are in both. Create a new column called `covid_per_capita` based on the total cases dividedby total population.

```{r aware_covid, error=TRUE}
# YOUR CODE HERE

aware_covid
```

```{r}
. = ottr::check("tests/aware_covid.R")
```

With all that done, I'll do the last few steps to make an interesting plot. Fortunately, the `covid_data` came with the classification of countries into continents, so we can add plenty of info to this plot!

```{r}
aware_covid |>
    pivot_longer(
        cols = starts_with("aware_"),
        names_prefix = "aware_",
        names_to = "awareness",
        values_to = "value"
    ) |>
    mutate(
        awareness = ordered(awareness,
            levels = c("refuse", "no", "alittle", "moderate", "alot")
        )
    ) |>
    ggplot() +
    aes(x = value, y = cases_per_capita, colour = continent) +
    geom_point() +
    geom_smooth(se = FALSE, method = "lm", formula = y ~ x) +
    facet_wrap(~ awareness, scales = "free") +
    labs(
        x = "Proportion of respondents who selected the awareness statement, by country",
        y = "Total COVID-19 cases per capita in 2020",
        title = "COVID-19 Cases versus Awareness of Climate Change Definitions",
        subtitle = "Facets represent responses to a question about climate change awareness.",
        colour = "Continent"
    )
```

Some interesting patterns:

- It looks like the "refused" group are pretty similar to the "no" group. It seems safe to say that the people who refuse to answer specific questions on a climate survey are not aware of the actual definitions being used.
- For the Americas, it looks like being more aware of climate change is associated with an *increase* in COVID-19 cases.
- For other continents, there doesn't appear to be any strong patterns.

There are several explanations for these patterns, such as awareness being linked to education and/or travel volume in first-world/developed countries, or perhaps it's just a consequence of how we define a "continent" (could you list the 18 countries in North America?). Most likely some of the results are due to the differences in the way countries track and report cases of COIVD-19.

# World Happiness Report Score (from 2023 report)

Finally, we have data from the [World Happiness Report](https://worldhappiness.report), downloaded from [here](https://worldhappiness.report/ed/2023/#appendices-and-data) and distributed with this assignment. The "Statistical Appendix" includes the description of each column, especially what the question was.

For the most part, the "life_ladder" column is the most important - this is the WHR's main question that gets at how "happy" the people of a country are. You may choose to explore other options though - perhaps people's perceptions of corruption affect their response to climate issues?

To make the data more convenient for Part 2, perform the following operations. The first ones are easy, the last one is tricky!

- Use `janitor` to clean the names. 
- Rename `country_name` to `country`.
- Remove any rows for which `life_ladder` is `NA`.
- The data contain information from multiple years, and these years are not consistent for each country. For Part 2, these data will mainly be used to compare countries (we don't need the happiness over time). Filter the data so that we only have the *most recent* observation of the relevant values. (In my solution, I use the `slice()` function as part of a very short pipeline.)

```{r happiness}
happiness <- read.xlsx("DataForTable2.1.xlsx") |>
    # YOUR CODE HERE

happiness
```
```{r}
. = ottr::check("tests/happiness.R")
```

And, finally, save the file:

```{r}
write_parquet(happiness, "happiness.parquet")
```

```{r active="", eval=FALSE}
# END ASSIGNMENT 
```

### 2024-11-01 A3-Pt1.Rmd
Update A3-Pt1.Rmd
at_cyclone All tests passed!

@@
 
-- 
-- 
-- 
-- 
--
+-   Aditya
+-   Andy
+-   Dibu
+-   Oscar
+-   Taras
 
@@
     ) |>
+    filter(!grepl("AL", `4`)) |>
     separate_wider_delim(
@@
 
-Write a function that takes in a vector of latitude _or_ longitude in the format of these data, then outputs a positive version of the numeric part if there's an "N" or an "E", and a negative version of the numeric part if there's a "W" or "S".
+Write a function that takes in a vector of latitude *or* longitude in the format of these data, then outputs a positive version of the numeric part if there's an "N" or an "E", and a negative version of the numeric part if there's a "W" or "S".
 
@@
 Categorize the cyclones based on the scale above using a `case_when()` statement based on the `max_wind` column.
-The code is set up to create an [*ordered factor*](https://fralfaro.github.io/r4ds/factors.html) so that R knows that the levels are ordered as TD < TS < 1 < 2 < 3 < 4 < 5.
+The code is set up to create an [*ordered factor*](https://fralfaro.github.io/r4ds/factors.html) so that R knows that the levels are ordered as TD \< TS \< 1 \< 2 \< 3 \< 4 \< 5.
+
 ```{r cyclones_data}
@@
 Now that we have a suitable data frame, save it as a parquet file with the file name `"cyclones_data.parquet"`.
+
 ```{r}
@@
 | year | region    | name | value  |
-|-------|-----------|------|---------|
+|------|-----------|------|--------|
 | 1978 | Antarctic | min  | 7.283  |
@@
 -   "Bolivia (Plurinational State of)", "Congo (Democratic Republic of the)", "Egypt", "Hong Kong", "Lao People's Democratic Republic", "Slovakia", "North MAcedonia", "South Korea" (be careful with this one!), "Turkey", "Vietnam", "Yemen".
+
 ```{r aware_fixed_country_match}
@@
 ```
+
 ```{r}

### 2024-11-01 A3-Pt1.Rmd
np_cyclone All tests passed!

@@
     ) |>
-    filter(!grepl("AL", `4`)) |>
     separate_wider_delim(
         cols = `4`,
-        # Set the delim and the names
-        # YOUR CODE HERE
+        delim = ",",
+        names = new_columns
     ) |>
@@
         # Specify the widths
-        # YOUR CODE HERE
+        widths = c(Basin = 2, Number = 2, NameYear = 4)
     ) |>
@@
         # Specify the widths
-        # YOUR CODE HERE
+        widths = c(ObservYear = 4, Month = 2, Day = 2)
     ) |>
@@
         # Specify the widths
-        # YOUR CODE HERE
+        widths = c(Hour = 2, Minute = 2)
     ) |>
@@
     # ALL of the steps all over again
-    # YOUR CODE HERE
+    read_csv(
+        col_names = c(as.character(1:4)),
+        progress = FALSE,
+        show_col_types = FALSE
+    ) |>
+    separate_wider_delim(
+        cols = `4`,
+        delim = ",",
+        names = new_columns
+    ) |>
+    mutate(
+        across(everything(), str_trim),
+        across(everything(), ~na_if(., "-999")),
+        across(everything(), ~na_if(., "-99")),
+        BasinNumberYear = ifelse(is.na(status), `1`, NA),
+        Name = ifelse(is.na(status), `2`, NA),
+        Entries = ifelse(is.na(status), `3`, NA)
+    ) |>
+    relocate(BasinNumberYear, Name, Entries) |>
+    fill(BasinNumberYear, Name, Entries) |>
+    filter(!is.na(status)) |>
+    select(-Entries) |>
+    separate_wider_position(
+        BasinNumberYear,
+        widths = c(Basin = 2, Number = 2, NameYear = 4)
+    ) |>
+    separate_wider_position(
+        `1`,
+        widths = c(ObservYear = 4, Month = 2, Day = 2)
+    ) |>
+    separate_wider_position(
+        `2`,
+        widths = c(Hour = 2, Minute = 2)
+    ) |>
+    rename(
+        Identifier = `3`
+    ) |>
+    mutate(
+        across(
+            c(NameYear, ObservYear, Month, Day, Hour, Minute, Number),
+            as.integer
+        )
+    ) |>
+    mutate(across(max_wind:r_max_wind, as.numeric))
 
 print(np_cyclone)
+
 ```

### 2024-11-03 A3-Pt1.Rmd
Update A3-Pt1.Rmd

@@
                 # YOUR CODE HERE
+                max_wind < 34 ~ "TD",
+                max_wind >= 34 & max_wind < 64 ~ "TS",
+                max_wind >= 64 & max_wind < 83 ~ "1",
+                max_wind >= 83 & max_wind < 96 ~ "2",
+                max_wind >= 96 & max_wind < 113 ~ "3",
+                max_wind >= 113 & max_wind < 137 ~ "4",
+                max_wind >= 137 ~ "5"
+
             ),
@@
 # YOUR CODE HERE
+write_parquet(cyclones_data, "cyclones_data.parquet")
 ```

### 2024-11-03 A3-Pt1.Rmd
cyclones_data_update_0 All tests passed!
Update A3-Pt1.Rmd

@@
 
--   Aditya
+-   Aditya (169027493)
 -   Andy
@@
 # YOUR CODE HERE
-
+cyclones_data_update_0 <- bind_rows(at_cyclone, np_cyclone)
 print(cyclones_data_update_0)
@@
     # YOUR CODE HERE
+    group_by(year, region) |>
+    summarize(
+        min = min(ice_extent, na.rm = TRUE),
+        max = max(ice_extent, na.rm = TRUE)
+    ) |>
+    pivot_longer(
+        cols = c("min", "max"),
+        names_to = "name",
+        values_to = "value"
+    )
 

### 2024-11-04 A3-Pt1.Rmd
Update A3-Pt1.Rmd

@@
         min = min(ice_extent, na.rm = TRUE),
-        max = max(ice_extent, na.rm = TRUE)
+        max = max(ice_extent, na.rm = TRUE),
+        .groups = "drop"
     ) |>
@@
     )
-
 ice_extent_yearly
@@
 # YOUR CODE HERE
+write_parquet(ice_extent_yearly, "ice_extent_yearly.parquet")
 ```

### 2024-11-04 A3-Pt1.Rmd
Update A3-Pt1.Rmd

@@
 title: "Exploring Disparate Data: Part 1"
-author: "Put your group number from MyLS here: "
+author: "Put your group number from MyLS here: 8"
 date: "Due November 12th"
@@
 convert_latlon <- function(latlon) {
-    # YOUR CODE HERE
+
+  parsed_num <- parse_number(latlon)
+    if_else(str_detect(latlon, "[SW]"), -parsed_num, parsed_num)
 }

### 2024-11-04 A3-Pt1.Rmd
489 - 672 done

--- 2024-11-03 A3-Pt1.Rmd
@@
 -   Andy
--   Dibu
+-   Dibu (169708451)
 -   Oscar
@@
     # YOUR CODE HERE
-    group_by(year, region) |>
-    summarize(
-        min = min(ice_extent, na.rm = TRUE),
-        max = max(ice_extent, na.rm = TRUE)
-    ) |>
-    pivot_longer(
-        cols = c("min", "max"),
-        names_to = "name",
-        values_to = "value"
-    )
 
@@
 ```{r covid_2020, error=TRUE}
-# YOUR CODE HERE
+covid_2020 <- covid_cases |>
+  filter(year(date) == 2020) |>
+  group_by(location, continent) |>
+  summarise(
+    total_cases = sum(total_cases)
+  ) |>
+  rename(country = location)
 
@@
 ```{r covid_2020_parquet, error=TRUE}
-# YOUR CODE HERE
+write_parquet(covid_2020, "covid_2020.parquet")
 ```
@@
         # Replace *all* *literal* periods with a space in the country column
-        # YOUR CODE HERE
+        country = str_replace_all(country, "\\.", " ")
     )
@@
       # Change country using a case_when statement
-        # YOUR CODE HERE
+      country = case_when(
+        country == "Bolivia (Plurinational State of)" ~ "Bolivia",
+        country == "Congo (Democratic Republic of the)" ~ "Congo, Rep.",
+        country == "Egypt" ~ "Egypt, Arab Rep.",
+        country == "Hong Kong" ~ "Hong Kong SAR, China",
+        country == "Lao People's Democratic Republic" ~ "Lao PDR",
+        country == "Slovakia" ~ "Slovak Republic",
+        country == "North MAcedonia" ~ "North Macedonia",
+        country == "South Korea" ~ "Korea, Rep.",
+        country == "Turkey" ~ "Turkiye",
+        country == "Vietnam" ~ "Viet Nam",
+        country == "Yemen" ~ "Yemen, Rep.",
+        TRUE ~ country
+      )
     )
@@
 ```{r aware_pop, error=TRUE}
-# YOUR CODE HERE
+
+aware_pop <- full_join(
+  aware_fixed_country,
+  pop_data,
+  by = "country"
+) |> full_join(population) |>
+  group_by(country, aware_no, aware_alittle, aware_moderate, aware_alot, aware_refuse) |>
+  summarise(total_population = sum(population))
 
@@
 ```{r covid_2020_match_country, error=TRUE}
-# YOUR CODE HERE
+covid_2020_match_country <- left_join(aware_pop, covid_2020, by = "country")
 

### 2024-11-04 A3-Pt1.Rmd
Update A3-Pt1.Rmd

@@
     mutate(
-        # YOUR CODE HERE
+        date = make_datetime(ObservYear, Month, Day, Hour * 100 + Minute)
     )
@@
     # YOUR CODE HERE
-
+    group_by(year, region) |>
+    summarize(
+        min = min(ice_extent, na.rm = TRUE),
+        max = max(ice_extent, na.rm = TRUE),
+        .groups = "drop"
+    ) |>
+    pivot_longer(
+        cols = c("min", "max"),
+        names_to = "name",
+        values_to = "value"
+    )
 ice_extent_yearly

### 2024-11-04 A3-Pt1.Rmd
Update A3-Pt1.Rmd

--- 2024-11-04 A3-Pt1.Rmd
@@
     mutate(
-        # YOUR CODE HERE
+        date = make_datetime(year = ObservYear,month = Month, day = Day, hour = Hour, min = Minute)
     )

### 2024-11-04 A3-Pt1.Rmd
Update A3-Pt1.Rmd
finished all tests passed

@@
 -   Dibu
--   Oscar
+-   Oscar (169057882)
 -   Taras
@@
 happiness <- read.xlsx("DataForTable2.1.xlsx") |>
-    # YOUR CODE HERE
+  janitor :: clean_names() |> 
+  rename(
+    country = country_name
+  ) |>
+  filter(!(is.na(life_ladder)))|>
+group_by(country) |>                 
+  arrange(country, desc(year)) |>      
+  slice(1)
 

### 2024-11-04 A3-Pt1.Rmd
Update A3-Pt1.Rmd
finished covid_2020

@@
 -   Oscar (169057882)
--   Taras
+-   Taras (169099616)
 
@@
 ```{r covid_2020, error=TRUE}
-# YOUR CODE HERE
+covid_2020 <- covid_cases |>
+
+  filter(year(date) == 2020) |>    
+  group_by(location) |>  
+  select(location, continent, total_cases)|>
+  rename(country = location)
 

### 2024-11-04 A3-Pt2.Rmd
Create A3-Pt2.Rmd

---
title: "TITLE OF YOUR PROJECT"
author: "PUT YOUR MYLS GROUP NUMBER HERE, THEN LIST GROUP MEMBERS BELOW"
date: "November 12th, 2024"
output: pdf_document
---

List your group members, including their student numbers, here:

- John K. Samson (#########)
- John Darnielle (#########)
- Craig Finn (#########)
- Joel Plaskett (########)
- Ezra Furman (#########)

You **must** be in a group in MyLS in order to see the DropBox used for submission. Even if you're alone, you must join a group by yourself.

You **must** be in a group with people from the same section as you. MyLS does not allow for groups including students from both Data100A and Data100B.

```{r setup, include=FALSE}
# echo = FALSE will set the Rmd to *not* show the R code. Don't change this.
# You may change the default figure width and figure height as you please.
knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.width = 6)

# Put any libraries that you need to load here.
# DO NOT PUT "install.packages()" IN AN RMD FILE!!!
library(tidyverse)
library(arrow)
```

# Instructions

You are encouraged to remove this instruction section prior to submission.

It is recommended that you follow the structure of this template. The text is all placeholder - you are free to change any/all wording as you please, but it is very helpful for the grading process if you keep the same structure. Anything in <<double angle brackets>> definitely needs to be changed, but you are free to change any/all sentences!

Note that all of the code is *hidden* by default. This file will be graded based on the insights, not the code.

You will only submit the PDF version of this document. To knit to PDF, you'll need to run `install.packages("tinytex")` in the console, followed by `tinytex::install_tinytex()` (DO NOT PUT THESE COMMANDS IN AN RMD FILE!!!). If you encounter errors in "Knit to PDF", you can "knit to html" and then print the html file to PDF using your operating system's PDF view (e.g. Adobe Acrobat). Only standalone PDF files will be accepted by MyLS.

# Abstract

General context, very brief data descriptions, techniques used, and general conclusions, all contained within a single, concise paragraph.

# Introduction

Climate change is something that has been studied. Here's some relevant information about the context of our study.

If needed, this paragraph is more information about the context.

In this report, we are going to explore some aspects climate change and the impact and/or perceptions of it by using exploratory techniques. We'll explore <<general description of data>> using <<general description of techniques>>.

By the end of this report, we will have shown ...

# Data Description

## <<Data Set 1>>

```{r load_data1}
# Put in your code to load in the data set, along with any
# necessary cleaning beyond what was done in Part 1

# Note that the code in this document will not be shown
# when you click "knit", so the placement of this code
# chunk is purely for your benefit: You can see what happened
# with your data, which makes it easier to describe below!
```

The data come from <<place>> and describe <<more specific description of the data>>.

In order to clean the data, we <<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code>>.

## <<Data Set 2>>

```{r load_data2}
# Put in your code to load in the data set, along with any
# necessary cleaning beyond what was done in Part 1

# Reminder: do NOT print your data to the screen unless it's
# completely necessary
```

The data come from <<place>> and detail <<more specific description of the data>>.

In order to clean the data, we <<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code>>

## <<Data Set 3>>

```{r load_data2}
# Put in your code to load in the data set, along with any
# necessary cleaning beyond what was done in Part 1
```

The data come from <<place>> and detail <<more specific description of the data>>.

In order to clean the data, we <<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code>>

## Combining the Data

Explain how any combinations of data were performed. Explain what kind of join was needed, whether columns had to be modified (for example, matching "country" names.)

# Exploratory Data Analysis

To achieve our goals, we explored the data by...

We explored many aspects of the data, but will demonstrate three. These are <<insight 1>>, <<insight 2>>, and <<insight3>>

The first aspect that we found interesting is shown in \@ref(fig:insight1). The insight should be specific to the data shown, not a general statement beyond the data (leave that for the conclusion).

```{r insight1, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
# This is an example of how you can control figures and captions in
# an R chunk. Note that you can reference figures using:
# \@ref(fig:insight1), where "insight1" is the label of this code
# chunk (the first bit of text after the "r" in "```{r label, options...}")
```

This insight is supported by the summary statistics in table \@ref(tab:summary_stats)

```{r summary_stats}
# Calculate the relevant summary statistics here.
# Note that the "kable" function in the "knitr" package
# is convenient for making nice tables. Other packages can
# do much fancier things with tables, but keep in mind that
# the insights should be the star, not the formatting.
```

The next insight that we found is shown in \@ref(fig:insight2).

```{r insight2, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
# This figure will have a height of 4 and a width of 6.
# Feel free to change this, and to apply different sizes
# to the other figures you create.
```

Finally, \@ref(fig:insight3) shows ...

```{r insight3, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
```

# Conclusion and Future Work

Overall, we found <<general ideas>>.

A second paragraph about our findings.

The next steps in this analysis are...

The limitations of this analysis are as follows. (Do not simply list potential issues with sampling, but relate them to your analysis and how they affect your conclusions. An honest and complete acknowledgement of the limitations makes the analysis more trustworthy.)

# References

I am not strict about MLA or APA style or anything like that. For this report, I would much rather have your citations be easy to match to your insights.

The easiest way is to use Rmd's [footnote](https://bookdown.org/yihui/rmarkdown/markdown-syntax.html#inline-formatting) syntax. This will put a number beside the word where the footnote appears, and the full text of the footnote at the bottom of the page (pdf) or end of the document (html). The syntax is:^[See the source view to see this footnote], where I suggest that you put in something like this^[The relevance to the insight is ... . From <<name of source and name of article>>, published on <<date>>, url: <<link to page>>] to make references for this assignment.

Alternatively, you could make a list of citations with their main arguments and why they're relevent to your insights, methods, etc.

The link above also references "bibtex" files. These are also extremely convenient, but have a steep learning curve and they make it difficult to tie them to an insight. If you use bibtext, then make sure that you provide a sentence to describe the source and it's relevance when you cite it - don't just add citations to the end of a sentence (this is common practice in academia, but I want to know that your citations are directly relevant for this assignmnet).

### 2024-11-04 A3-Pt2.Rmd
Update A3-Pt2.Rmd

@@
 title: "TITLE OF YOUR PROJECT"
-author: "PUT YOUR MYLS GROUP NUMBER HERE, THEN LIST GROUP MEMBERS BELOW"
+author: "Group 8"
 date: "November 12th, 2024"
@@
 
-- John K. Samson (#########)
+- Aditya Chauhan (169027493)
 - John Darnielle (#########)

### 2024-11-04 A3-Pt2.Rmd
load database

--- 2024-11-04 A3-Pt2.Rmd
@@
 
-- John K. Samson (#########)
-- John Darnielle (#########)
-- Craig Finn (#########)
-- Joel Plaskett (########)
-- Ezra Furman (#########)
+-   John K. Samson (\#########)
+-   John Darnielle (\#########)
+-   Craig Finn (\#########)
+-   Joel Plaskett (\########)
+-   Ezra Furman (\#########)
 
@@
 The text is all placeholder - you are free to change any/all wording as you please, but it is very helpful for the grading process if you keep the same structure.
-Anything in <<double angle brackets>> definitely needs to be changed, but you are free to change any/all sentences!
+Anything in \<<double angle brackets>\> definitely needs to be changed, but you are free to change any/all sentences!
 
@@
 In this report, we are going to explore some aspects climate change and the impact and/or perceptions of it by using exploratory techniques.
-We'll explore <<general description of data>> using <<general description of techniques>>.
+We'll explore \<<general description of data>\> using \<<general description of techniques>\>.
 
@@
 
-## <<Data Set 1>>
+## \<\<Data Set 1\>\>
 
@@
 # necessary cleaning beyond what was done in Part 1
-
+cyclones_data <- read_parquet("cyclones_data.parquet")
+cyclones_data
 # Note that the code in this document will not be shown
@@
 
-The data come from <<place>> and describe <<more specific description of the data>>.
+The data come from \<<place>\> and describe \<<more specific description of the data>\>.
 
-In order to clean the data, we <<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code>>.
+In order to clean the data, we \<\<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code\>\>.
 
-## <<Data Set 2>>
+## \<\<Data Set 2\>\>
 
@@
 # necessary cleaning beyond what was done in Part 1
-
+ice_extent_yearly <- read_parquet("ice_extent_yearly.parquet")
+ice_extent_yearly
 # Reminder: do NOT print your data to the screen unless it's
@@
 
-The data come from <<place>> and detail <<more specific description of the data>>.
+The data come from \<<place>\> and detail \<<more specific description of the data>\>.
 
-In order to clean the data, we <<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code>>
+In order to clean the data, we \<\<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code\>\>
 
-## <<Data Set 3>>
+## \<\<Data Set 3\>\>
 
@@
 # necessary cleaning beyond what was done in Part 1
+climate_awareness <- read_parquet("climate_awareness.parquet")
+climate_awareness
 ```
 
-The data come from <<place>> and detail <<more specific description of the data>>.
+The data come from \<<place>\> and detail \<<more specific description of the data>\>.
 
-In order to clean the data, we <<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code>>
+In order to clean the data, we \<\<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code\>\>
 
@@
 
-We explored many aspects of the data, but will demonstrate three. These are <<insight 1>>, <<insight 2>>, and <<insight3>>
+We explored many aspects of the data, but will demonstrate three. These are \<\<insight 1\>\>, \<\<insight 2\>\>, and \<<insight3>\>
 
@@
 
-Overall, we found <<general ideas>>.
+Overall, we found \<<general ideas>\>.
 
@@
 This will put a number beside the word where the footnote appears, and the full text of the footnote at the bottom of the page (pdf) or end of the document (html).
-The syntax is:^[See the source view to see this footnote], where I suggest that you put in something like this^[The relevance to the insight is ...
-.
-From <<name of source and name of article>>, published on <<date>>, url: <<link to page>>] to make references for this assignment.
+The syntax is:[^1], where I suggest that you put in something like this[^2] to make references for this assignment.
+
+[^1]: See the source view to see this footnote
+
+[^2]: The relevance to the insight is ... . From \<<name of source and name of article>\>, published on \<<date>\>, url: \<<link to page>\>
 

### 2024-11-04 A3-Pt1.Rmd
covid and world happiness

--- 2024-11-04 A3-Pt1.Rmd
@@
     mutate(
-        date = make_datetime(year = ObservYear,month = Month, day = Day, hour = Hour, min = Minute)
+        date = make_datetime(ObservYear, Month, Day, Hour * 100 + Minute)
     )
@@
 ```{r aware_pop, error=TRUE}
-# YOUR CODE HERE
+
+aware_pop <- inner_join(
+  aware_fixed_country_match,
+  pop_data,
+  by = "country"
+) |> summarise(country, "total_population" = SP.POP.TOTL, aware_no, aware_alittle, aware_moderate, aware_alot, aware_refuse)
 
@@
 ```{r aware_covid, error=TRUE}
-# YOUR CODE HERE
+aware_covid <- inner_join(aware_pop, covid_2020_match_country) |>
+  group_by(country) |>
+  mutate(
+    "cases_per_capita" = total_cases / total_population
+  )
 
@@
 happiness <- read.xlsx("DataForTable2.1.xlsx") |>
-    # YOUR CODE HERE
+  clean_names() |>
+  rename(
+    "country" = country_name
+  ) |>
+  filter(
+    !is.na(life_ladder)
+  ) |>
+  group_by(
+    country
+  ) |>
+  slice_max(
+    order_by = year
+  )
 

### 2024-11-04 A3-Pt1.Rmd

@@
 
--   Aditya (169027493)
--   Andy
+-   Andy (169081356)
 -   Dibu (169708451)
--   Oscar
--   Taras
+-   Oscar (169057882)
+-   Taras (169099616)
 
@@
 ```{r error=TRUE}
-# YOUR CODE HERE
 cyclones_data_update_0 <- bind_rows(at_cyclone, np_cyclone)
@@
     mutate(
-        date = make_datetime(ObservYear, Month, Day, Hour * 100 + Minute)
+        date = make_datetime(year = ObservYear,month = Month, day = Day, hour = Hour, min = Minute)
     )
@@
             case_when(
-                # YOUR CODE HERE
                 max_wind < 34 ~ "TD",
@@
 ```{r}
-# YOUR CODE HERE
 write_parquet(cyclones_data, "cyclones_data.parquet")
@@
 ice_extent_yearly <- ice_extent_daily |>
-    # YOUR CODE HERE
     group_by(year, region) |>
@@
 ```{r sea_ice_parquet, error=TRUE}
-# YOUR CODE HERE
 write_parquet(ice_extent_yearly, "ice_extent_yearly.parquet")
@@
 happiness <- read.xlsx("DataForTable2.1.xlsx") |>
-  clean_names() |>
+  janitor :: clean_names() |> 
   rename(
-    "country" = country_name
+    country = country_name
   ) |>
-  filter(
-    !is.na(life_ladder)
-  ) |>
-  group_by(
-    country
-  ) |>
-  slice_max(
-    order_by = year
-  )
+  filter(!(is.na(life_ladder)))|>
+group_by(country) |>                 
+  arrange(country, desc(year)) |>      
+  slice(1)
 

### 2024-11-14 Draft.Rmd
insight 2

---
title: "Analysis of Global Crises and Their Impact on Interational Well-Being"
author: "Group 8"
date: "November 12, 2024"
output: pdf_document
---

List your group members, including their student numbers, here:

- Aditya Chauhan (169027493)
- Andy Ferrer Lorenzo (169081356)
- Oscar Saul (169057882)
- Taras Zakydalsky (169099616)
- Dibuhang Rai (169708451)

You **must** be in a group in MyLS in order to see the DropBox used for submission. Even if you're alone, you must join a group by yourself.

You **must** be in a group with people from the same section as you. MyLS does not allow for groups including students from both Data100A and Data100B.

```{r initial_setup, include=FALSE}
# echo = FALSE will set the Rmd to *not* show the R code. Don't change this.
# You may change the default figure width and figure height as you please.
knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.width = 6)
# Put any libraries that you need to load here.
# DO NOT PUT "install.packages()" IN AN RMD FILE!!!
library(tidyverse)
library(arrow)
library(maps)
library(tinytex)
library(sf)
library(rnaturalearth)
library(corrplot)
library(scales)
library(geosphere)
library(ggridges)
```

# Instructions

You are encouraged to remove this instruction section prior to submission.

It is recommended that you follow the structure of this template. The text is all placeholder - you are free to change any/all wording as you please, but it is very helpful for the grading process if you keep the same structure. Anything in \<<double angle brackets>\> definitely needs to be changed, but you are free to change any/all sentences!

Note that all of the code is *hidden* by default. This file will be graded based on the insights, not the code.

You will only submit the PDF version of this document. To knit to PDF, you'll need to run `install.packages("tinytex")` in the console, followed by `tinytex::install_tinytex()` (DO NOT PUT THESE COMMANDS IN AN RMD FILE!!!). If you encounter errors in "Knit to PDF", you can "knit to html" and then print the html file to PDF using your operating system's PDF view (e.g. Adobe Acrobat). Only standalone PDF files will be accepted by MyLS.

# Purpose
This study aims to

# Abstract

# Introduction

Climate change is something that has been studied. Here's some relevant information about the context of our study.

If needed, this paragraph is more information about the context.

In this report, we are going to explore some aspects climate change and the impact and/or perceptions of it by using exploratory techniques. We'll explore \<<general description of data>\> using \<<general description of techniques>\>.

By the end of this report, we will have shown ...

# Data Description

## \<\<Cyclones Data\>\>
```{r load_data1}
cyclones_data <- read_parquet("cyclones_data.parquet")
world <- ne_countries(scale = "medium", returnclass = "sf")

# ________________Map long and lat to scale________________
# Convert latitude(-90/+90) and longitude(-180/+180) to be within the correct range
cyclones_data <- cyclones_data |>
    rename(long = lon) |> #rename lon to long to match with world map
    mutate(
        lat = ((lat + 90) %% 180) - 90,  
        long = ((long + 180) %% 360) - 180 
    )

# ________________Determine distance of cyclone to equator________________
equator_point <- c(0, 0)
cyclones_data <- cyclones_data %>%
  mutate(
    km_distance_from_equator = distGeo(cbind(long, lat), cbind(long, 0)) / 1000
  )

# ________________Determine distance of cyclone to equator________________
equator_point <- c(0, 0)
cyclones_data <- cyclones_data %>%
  mutate(
    km_distance_from_equator = distGeo(cbind(long, lat), cbind(long, 0)) / 1000
  )

# ________________Convert long and lat points to countries________________
cyclones_data <- st_as_sf(cyclones_data, coords = c("long", "lat"), crs = 4326, remove = FALSE) |>
  st_join(world["name"], left = TRUE) |>
  rename(country = name)

# ________________Create cyclone_cleaned with specified columns________________
cyclone_cleaned <- cyclones_data |>
    select(country, lat, long, category, km_distance_from_equator)

cyclone_cleaned

```

The data set `cyclone_data` provided by the National Hurricane Center and the Central Pacific Hurricane Center records hurricane entries in the Atlantic and North Pacific basins from 1851 to 2022 and 1949 to 2022, respectively. The data frame includes detailed time stamps, coordinates, and categories, which are vital for analysis.

The preliminary step in supporting the analysis involved converting the coordinates into the Cartesian coordinate system. Next, we used the `rnaturalearth` package and `geolocation` to convert the coordinates into country identifiers and determine each storm's distance in kilometers from the equator. As shown in Figure 1 in the appendix, cyclones are now accurately distributed across the North Atlantic and Northeast Pacific basins with the correct identifiers.

## \<\<World Covid 2020\>\>

```{r load_data2}
covid_2020 <- read_parquet("covid_2020.parquet")
world <- ne_countries(scale = "medium", returnclass = "sf")

covid_2020 <- covid_2020 %>%
  mutate(country = case_when(
  country == "British Virgin Islands" ~ "British Virgin Is.",
  country == "Cape Verde" ~ "Cabo Verde",
  country == "Cayman Islands" ~ "Cayman Is.",
  country == "Central African Republic" ~ "Central African Rep.",
  country == "Cook Islands" ~ "Cook Is.",
  country == "Cote d'Ivoire" ~ "Ivory Coast",
  country == "Democratic Republic of Congo" ~ "Congo",
  country == "Cyprus" ~ "Cyprus",
  country == "Dominican Republic" ~ "Dominican Rep.",
  country == "East Timor" ~ "Timor-Leste",
  country == "England" ~ "United Kingdom",
  country == "Equatorial Guinea" ~ "Equat. Guinea",
  country == "Eswatini" ~ "Swaziland",
  country == "Falkland Islands" ~ "Falkland Isles",
  country == "Faroe Islands" ~ "Faroe Isles",
  country == "French Guiana" ~ "Fr. Guiana",
  country == "French Polynesia" ~ "Fr. Polynesia",
  country == "Gibraltar" ~ "Gibraltar",
  country == "Guadeloupe" ~ "Guadeloupe",
  country == "Marshall Islands" ~ "Marshall Is.",
  country == "Martinique" ~ "Martinique",
  country == "Mayotte" ~ "Mayotte",
  country == "Micronesia (country)" ~ "Micronesia",
  country == "Northern Ireland" ~ "Northern Ireland",
  country == "Northern Mariana Islands" ~ "N. Mariana Is.",
  country == "Pitcairn Islands" ~ "Pitcairn",
  country == "Reunion" ~ "Réunion",
  country == "Saint Barthelemy" ~ "Saint-Barthélemy",
  country == "Saint Kitts and Nevis" ~ "St Kitts & Nevis",
  country == "Saint Martin (French part)" ~ "Saint Martin",
  country == "Saint Pierre and Miquelon" ~ "St Pierre & Miquelon",
  country == "Saint Vincent and the Grenadines" ~ "St Vincent & Gren.",
  country == "Sao Tome and Principe" ~ "São Tomé",
  country == "Scotland" ~ "Scotland",
  country == "Sint Maarten (Dutch part)" ~ "Sint Maarten",
  country == "Solomon Islands" ~ "Solomon Is.",
  country == "South Sudan" ~ "S. Sudan",
  country == "Tokelau" ~ "Tokelau",
  country == "Turks and Caicos Islands" ~ "Turks & Caicos",
  country == "United States Virgin Islands" ~ "US Virgin Is.",
  country == "Wales" ~ "Wales",
  country == "Wallis and Futuna" ~ "Wallis & Futuna",
  country == "United States" ~ "United States of America",
  TRUE ~ country  
))

equator <- st_sfc(st_linestring(rbind(c(-180, 0), c(180, 0))), crs = st_crs(world))
world <- st_transform(world, crs = 3857)
equator <- st_transform(equator, crs = 3857)

world_with_distance <- world %>%
  mutate(
    centroid = st_centroid(geometry),
    latitude = st_coordinates(centroid)[, 2],
    hemisphere = ifelse(latitude > 0, "North", "South"),
    distance_to_equator_km = as.numeric(st_distance(centroid, equator)) / 1000  # Convert to kilometers
  ) %>%
  select(name, hemisphere, distance_to_equator_km) %>%
  rename(country = name)

covid_2020 <- covid_2020 %>%
  left_join(world_with_distance, by = "country")

covid_2020

```

The `covid_2020` data set, provided by Our World in Data contains a comprehensive list of the total reported cases of COVID-19 cases across many countries and continents in 2020. To aid in the analysis `rnaturalearth` and `geolocation` was used to find the hemisphere and distance to the equator in kilometers.

## \<\<World Happiness Census\>\>

```{r load_data3}
# Put in your code to load in the data set, along with any
# necessary cleaning beyond what was done in Part 1
happiness <- read_parquet("happiness.parquet")
world_map <- map_data("world")

happiness_mapped <- happiness |>
    mutate(country = case_when( #match country names to world_map
        country == "United States" ~ "United States of America",
        country == "United Kingdom" ~ "UK",
        country == "Congo (Brazzaville)" ~ "Congo",
        country == "Congo (Kinshasa)" ~ "Congo",
        country == "Hong Kong S.A.R. of China" ~ "Hong Kong",
        country == "Türkiye" ~ "Turkey",
        country == "Lao.People's.Democratic.Republic" ~ "Laos",
        country == "United Arab Emirates" ~ "UAE",
        TRUE ~ country
    )) |>
    right_join(world_map, by = c("country" = "region")) #joins the long and lat of the respective countries to map them

world <- ne_countries(scale = "medium", returnclass = "sf")

equator <- st_sfc(st_linestring(rbind(c(-180, 0), c(180, 0))), crs = st_crs(world))
world <- st_transform(world, crs = 3857)
equator <- st_transform(equator, crs = 3857)

happiness <- happiness |>
  left_join(
    world |>
      mutate(
        centroid = st_centroid(geometry),
        latitude = st_coordinates(centroid)[, 2],
        hemisphere = ifelse(latitude > 0, "North", "South"),
        distance_to_equator_km = as.numeric(st_distance(centroid, equator)) / 1000
      ) |>
      select(name, hemisphere, distance_to_equator_km) |>
      rename(country = name),
    by = "country"
  )

```

`happiness` dataset is global census on a countries economic, social, and political stability.

## Combining the Data

Explain how any combinations of data were performed. Explain what kind of join was needed, whether columns had to be modified (for example, matching "country" names.)

# Exploratory Data Analysis

We explored many aspects of the data, but will demonstrate three. These are \<\<Geographic and environmental Factors VS a countries overall well being\>\>, \<\<Covid Cases Vs perception of corruption\>\>, and \<<Freedom to Make Life Choices & Climate Awareness (OR DISTANCE VS COVID)>\>

The analysis highlighted an interesting finding, as shown in @ref(fig
): a significant concentration of COVID-19 cases appears within the 1500-2500 km range from the equator, suggesting a strong relationship between COVID-19 cases and cyclone activity within a margin of approximately 9 degrees of longitude. This area also highlights a geographical strip pointing to a genreal trend of lower GDP, government stability, and overall happiness.

```{r insight1, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}

covid_2020_world <- covid_2020 %>%
  filter(
    total_cases < 50000000 
  )

ggplot(covid_2020_world, aes(x = distance_to_equator_km, y = total_cases)) +
  geom_point(color = "steelblue", alpha = 0.3) +  # Lower alpha for more transparency
  geom_density_2d(color = "darkorange") +  # Density lines to show clustering
  geom_vline(xintercept = 1500, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 2500, linetype = "dashed", color = "black")  +
  labs(
    title = "Scatter Plot of COVID-19 Cases by Distance from Equator",
    x = "Distance from Equator (km)",
    y = "Total COVID-19 Cases"
  ) +
  coord_cartesian(xlim = c(0, 10250)) +
  scale_y_continuous(
    labels = scales::comma,
    trans = "log10",
    breaks = c(1, 10, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7, 5e7)  
  ) +
  theme_minimal()

ggplot(cyclones_data |> filter(!is.na(category)), aes(x = km_distance_from_equator, y = category, fill = category)) +
  geom_density_ridges(scale = 2, alpha = 0.7, color = "black") +
  geom_vline(xintercept = 1500, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 2500, linetype = "dashed", color = "black") +
  labs(
    x = "Distance from Equator (km)",
    y = "Cyclone Category",
    title = "Ridgeline Plot of Distance from Equator by Cyclone Category"
  ) +
  coord_cartesian(xlim = c(0, 6500)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10)
  )

```

A report conducted by the [National Librarbty of Medicine](https://pmc.ncbi.nlm.nih.gov/articles/PMC8079387/) highlights a strong correlation between COVID-19 cases and proximity to the equator, suggesting that certain geographic and environmental factors may influence virus spread and outbreak. This observation signifies a trend where regions near the equator are subject to unique vulnerabilities, not only experiencing increased COVID-19 cases but also facing intensified cyclone activity (as shown in the figure above), volcanic activity, elevated ocean currents, seasonal flooding, and the presence of natural carbon sinks.

Applying this insight into the analysis, it is evident that there are strong correlations to geological crises such as cyclones and COVID-19, and key indicators like GDP, life satisfaction, social support, and the overall well-being of countries.These crises profoundly impact not only economic stability but also the social and health of a nation, as evidenced by happiness levels in Figure 5 of the appendix. This reinforces the insight that crisis have a strong effect on the affected regions economic and social dynamics.

```{r}
#Social support & life ladder

data_combined <- inner_join(covid_2020, happiness, by = "country")

median_social_support <- median(data_combined$social_support, na.rm = TRUE)

ggplot(data_combined, aes(x = social_support, y = life_ladder, color = continent)) +
  geom_point(alpha = 0.7, position = "jitter") + 
  geom_smooth(method = "lm", color = "blue", se = TRUE) + 
  geom_vline(xintercept = median_social_support, color = "red", linetype = "dashed", linewidth = 1) + 
  labs(
    title = "Comparison of Social Support and Life Ladder by Continent",
    x = "Social Support",
    y = "Life Ladder (Happiness Score)",
    color = "Continent"
  ) +
  facet_wrap(~ continent) + 
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "grey", linetype = "dashed", linewidth = 0.5))

# Social Support & Life Expectancy

ggplot(data_combined, aes(x = social_support, y = healthy_life_expectancy_at_birth, color = continent)) +
  geom_point(alpha = 0.7, position = "jitter") + 
  geom_smooth(method = "lm", color = "blue", se = TRUE) + 
  geom_vline(xintercept = median_social_support, color = "red", linetype = "dashed", linewidth = 1) + 
  labs(
    title = "Comparison of Social Support and Life Expectancy by Continent",
    x = "Social Support",
    y = "Healthy Life Expectancy at Birth",
    color = "Continent"
  ) +
  facet_wrap(~ continent) + 
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "grey", linetype = "dashed", linewidth = 0.5))

data_combined <- inner_join(covid_2020, happiness, by = "country")

median_total_cases <- median(data_combined$total_cases, na.rm = TRUE)

ggplot(data_combined, aes(x = total_cases / 100, y = freedom_to_make_life_choices)) +
  geom_point(alpha = 0.5, position = "jitter") + 
  geom_smooth(method = "lm", color = "blue", se = TRUE) + 
  geom_vline(xintercept = median_total_cases, color = "red", linetype = "dashed", linewidth = 1) + 
  labs(
    title = "Comparison of COVID-19 Cases and Freedom",
    x = "Total COVID-19 Cases",
    y = "Freedom"
  ) +
  scale_x_log10() +  
  scale_y_log10() + 
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "grey", linetype = "dashed", linewidth = 0.5))

#Life expectancy & Cyclone count

cyclone_counts <- cyclones_data |>
  group_by(country) |>
  summarise(cyclone_count = n())

combined_data <- cyclone_counts |>
  inner_join(happiness, by = "country") |>
  select(country, cyclone_count, healthy_life_expectancy_at_birth)

combined_data <- combined_data |>
  arrange(desc(cyclone_count)) |>
  slice(-1:-3)

ggplot(combined_data, aes(y = cyclone_count, x = healthy_life_expectancy_at_birth)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Relationship Between Cyclone Count and Life Expectancy (Outlier Removed)",
    x = "Cyclone Count",
    y = "Healthy Life Expectancy at Birth"
  ) +
  theme_minimal()

colnames(happiness)

# distance & cyclone count

# Load required libraries
library(tidyverse)

# Step 1: Aggregate cyclone counts by country
cyclone_counts <- cyclones_data %>%
  group_by(country) %>%
  summarise(cyclone_count = n())

# Step 2: Join with happiness data on 'country' to add distance from the equator
combined_data <- cyclone_counts %>%
  inner_join(happiness, by = "country") %>%
  select(country, cyclone_count, distance_to_equator_km)

# Step 3: Remove the top 3 highest cyclone counts
combined_data <- combined_data %>%
  arrange(desc(cyclone_count)) %>%
  slice(-1:-3)

# Step 4: Plot the relationship between distance to equator and cyclone count
ggplot(combined_data, aes(x = distance_to_equator_km, y = cyclone_count)) +
  geom_point(alpha = 0.7, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Comparison of Cyclone Count and Distance from Equator (Top 3 Outliers Removed)",
    x = "Distance from Equator (km)",
    y = "Cyclone Count"
  ) +
  theme_minimal()

ggplot(cyclones_data %>% filter(!is.na(category)), aes(x = km_distance_from_equator, y = category, fill = category)) +
  geom_density_ridges(scale = 2, alpha = 0.7, color = "black", size = 0.3) +
  labs(
    x = "Distance from Equator (km)",
    y = "Cyclone Category",
    title = "Ridgeline Plot of Distance from Equator by Cyclone Category"
  ) +
  coord_cartesian(xlim = c(0, 6500)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10)
  )

colnames(happiness)

ggplot(cyclones_data |> filter(!is.na(category)), aes(x = km_distance_from_equator, y = category, fill = category)) +
  geom_density_ridges(scale = 2, alpha = 0.7, color = "black") +
  annotate("rect", xmin = 1500, xmax = 2500, ymin = 0, ymax = Inf, fill = "orange", alpha = 0.8) +
  labs(
    x = "Distance from Equator (km)",
    y = "Cyclone Category",
    title = "Ridgeline Plot of Distance from Equator by Cyclone Category"
  ) +
  coord_cartesian(xlim = c(0, 6500)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10)
  )

happiness <- happiness|>
  left_join(
    rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") |>
      transmute(
        country = name_long,
        distance_to_equator_km = geosphere::distGeo(
          cbind(st_coordinates(st_centroid(geometry))[,1], st_coordinates(st_centroid(geometry))[,2]),
          cbind(st_coordinates(st_centroid(geometry))[,1], 0)
        ) / 1000  # Convert to km
      ),
    by = "country"
  )

happiness

happiness_bands_gdp <- happiness |>
  mutate(distance_band = cut(distance_to_equator_km.y, breaks = seq(0, 10000, by = 500), include.lowest = TRUE, labels = seq(0, 9500, by = 500))) |>
  group_by(distance_band) |>
  summarize(avg_log_gdp_per_capita = mean(log_gdp_per_capita, na.rm = TRUE))

# Plot for log GDP per capita
ggplot(happiness_bands_gdp, aes(x = distance_band, y = avg_log_gdp_per_capita)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  annotate("rect", xmin = "1500", xmax = "2500", ymin = 0, ymax = Inf, fill = "orange", alpha = 0.8) +
  labs(
    title = "Average GDP per Capita by Distance from Equator",
    x = "Distance from Equator (km, Binned)",
    y = "Average Log GDP per Capita"
  ) +
  scale_x_discrete(breaks = seq(0, 9500, by = 1000)) +
  coord_cartesian(ylim = c(min(happiness_bands_gdp$avg_log_gdp_per_capita, na.rm = TRUE) - 0.1, 
                           max(happiness_bands_gdp$avg_log_gdp_per_capita, na.rm = TRUE) + 0.1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

# Group data into 500 km distance bands and calculate average life ladder score for each band
happiness_bands_life <- happiness |>
  mutate(distance_band = cut(distance_to_equator_km.y, breaks = seq(0, 10000, by = 500), include.lowest = TRUE, labels = seq(0, 9500, by = 500))) %>%
  group_by(distance_band) |>
  summarize(avg_life_ladder = mean(life_ladder, na.rm = TRUE))

# Plot for life ladder
ggplot(happiness_bands_life, aes(x = distance_band, y = avg_life_ladder)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  annotate("rect", xmin = "1500", xmax = "2500", ymin = 0, ymax = Inf, fill = "orange", alpha = 0.8) +
  labs(
    title = "Average Life Ladder Score by Distance from Equator",
    x = "Distance from Equator (km, Binned)",
    y = "Average Life Ladder Score"
  ) +
  scale_x_discrete(breaks = seq(0, 9500, by = 1000)) +
  coord_cartesian(ylim = c(min(happiness_bands_life$avg_life_ladder, na.rm = TRUE) - 0.1, 
                           max(happiness_bands_life$avg_life_ladder, na.rm = TRUE) + 0.1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```

This insight is supported by the summary statistics in table \@ref(tab:summary_stats)

The next insight that we found is shown in \@ref(fig:insight2).

```{r insight3, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}

covid_happiness <- inner_join(covid_2020_world, happiness, by = "country")

filter_covid_happiness <- covid_happiness |> filter(!is.na(total_cases), !is.na(perceptions_of_corruption))

ratio_covid_happiness <- filter_covid_happiness |> group_by(country) |>
  mutate(
    ratio = total_cases * perceptions_of_corruption
  )

covid_happiness |> filter(!is.na(total_cases), !is.na(perceptions_of_corruption)) |>
  ggplot(mapping = aes(x = perceptions_of_corruption, y = total_cases, color = continent)) + geom_boxplot() + facet_wrap(~continent) + labs(title = "Perception of Corruption and COVID cases by continent", subtitle = "Weight of COVID cases repective to perception of corruption", x = "Perception of Corruption (0 - 1)", y = "Total Cases", caption = "Source: covid.ourworldindata.org\nworldhappiness.report") + theme_minimal() + theme(legend.position = "None")

```

Finally, \@ref(fig:insight3) shows ...

```{r insight3, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}

```

# Conclusion and Future Work

Overall, we found \<<general ideas>\>.

A second paragraph about our findings.

The next steps in this analysis are...

The limitations of this analysis are as follows. (Do not simply list potential issues with sampling, but relate them to your analysis and how they affect your conclusions. An honest and complete acknowledgement of the limitations makes the analysis more trustworthy.)

# References

I am not strict about MLA or APA style or anything like that. For this report, I would much rather have your citations be easy to match to your insights.

The easiest way is to use Rmd's [footnote](https://bookdown.org/yihui/rmarkdown/markdown-syntax.html#inline-formatting) syntax. This will put a number beside the word where the footnote appears, and the full text of the footnote at the bottom of the page (pdf) or end of the document (html). The syntax is:[^1], where I suggest that you put in something like this[^2] to make references for this assignment.

[^1]: See the source view to see this footnote

[^2]: The relevance to the insight is ... . From \<<name of source and name of article>\>, published on \<<date>\>, url: \<<link to page>\>

Alternatively, you could make a list of citations with their main arguments and why they're relevent to your insights, methods, etc.

The link above also references "bibtex" files. These are also extremely convenient, but have a steep learning curve and they make it difficult to tie them to an insight. If you use bibtext, then make sure that you provide a sentence to describe the source and it's relevance when you cite it - don't just add citations to the end of a sentence (this is common practice in academia, but I want to know that your citations are directly relevant for this assignmnet).

# Apendix

```{r world_map_cyclones_by_category, fig.cap="North Atlantic and Northeast Pacific Cyclones by Category"}
world_map <- map_data("world")

cyclones_data <- cyclones_data |>
    mutate(MonthName = factor(month.abb[Month], levels = month.abb)) |>
    relocate(MonthName, .after = Month)

ggplot() +
    geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "lightgray", color = "white") +
    geom_rect(aes(xmin = -180, xmax = 180, ymin = -5, ymax = 5), fill = "red", alpha = 0.2) + 
    geom_hex(data = cyclones_data, aes(x = long, y = lat, fill = category), bins = 50, alpha = 0.7) +
    scale_fill_viridis_d(name = "Storm Category") +
    labs(
        title = "North Atlantic and Northeast Pacific Cyclones by Category",
        x = "Longitude",
        y = "Latitude"
    ) +
    scale_x_continuous(breaks = c(-180, -90, 0, 90, 180)) +
    scale_y_continuous(breaks = c(-90, -45, 0, 45, 90)) +
    theme_minimal()
```

```{r world_map_cyclones_by_basin, fig.cap="North Atlantic and Northeast Pacific Cyclones by Basin"}
ggplot() +
    geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "lightgray", color = "white") +
    geom_rect(aes(xmin = -180, xmax = 180, ymin = -5, ymax = 5), fill = "red", alpha = 0.2) +
    geom_hex(data = cyclones_data, aes(x = long, y = lat, fill = Basin, alpha = ..count..), bins = 50) +
    scale_fill_viridis_d(name = "Storm Category") +
    scale_alpha(range = c(0.35, 1.5)) +
    labs(
        title = "North Atlantic and Northeast Pacific Cyclones by Basin",
        x = "Longitude",
        y = "Latitude"
    ) +
    scale_x_continuous(breaks = c(-180, -90, 0, 90, 180)) +
    scale_y_continuous(breaks = c(-90, -45, 0, 45, 90)) +
    theme_minimal()
```

```{r storm_frequency_by_month_year, fig.cap="Storm Frequency by Month and Year"}
ggplot(cyclones_data) +
    aes(x = ObservYear, y = MonthName, fill = ..count..) +
    geom_bin2d() +
    scale_fill_distiller(direction = 1, palette = "YlOrRd") +
    labs(
        title = "Storm Frequency by Month and Year",
        x = "Observation Year",
        y = "Month",
        fill = "Frequency"
    ) +
    theme_minimal()
```

```{r proportion_storm_categories_over_time, fig.cap="Proportion of Storm Categories Over Time"}
ggplot(cyclones_data) +
    aes(x = ObservYear, fill = category) +
    geom_bar(position = "fill", width = 1) +
    labs(
        title = "Proportion of Storm Categories Over Time",
        x = "Observation Year",
        y = "Proportion",
        fill = "Storm Category"
    ) +
    theme_minimal()
```

```{r world_happiness_scores, fig.cap="World Happiness Scores by Country"}
happiness_colors <- c("1" = "#000000", "3" = "#aa0000", "4" = "#f46d43",
                      "5" = "#fdae61", "6" = "#fee08b", "7" = "#66bd63",
                      "8" = "#1a9850")

ggplot(happiness_mapped) +
    aes(x = long, y = lat, group = group, fill = factor(round(life_ladder))) +
    geom_polygon(color = "white") +
    scale_fill_manual(values = happiness_colors, name = "Happiness\nScore") +
    theme_void() +
    labs(title = "World Happiness Scores by Country")
```

### 2024-11-14 Draft.Rmd

@@
 equator_point <- c(0, 0)
-cyclones_data <- cyclones_data %>%
+cyclones_data <- cyclones_data |>
   mutate(
@@
 
-covid_2020 <- covid_2020 %>%
+covid_2020 <- covid_2020 |>
   mutate(country = case_when(
@@
 
-world_with_distance <- world %>%
+world_with_distance <- world |>
   mutate(
@@
     distance_to_equator_km = as.numeric(st_distance(centroid, equator)) / 1000  # Convert to kilometers
-  ) %>%
-  select(name, hemisphere, distance_to_equator_km) %>%
+  ) |>
+  select(name, hemisphere, distance_to_equator_km) |>
   rename(country = name)
 
-covid_2020 <- covid_2020 %>%
+covid_2020 <- covid_2020 |>
   left_join(world_with_distance, by = "country")
@@
 
-`happiness` dataset is global census on a countries economic, social, and political stability.
+`happiness` dataset is global census on a countries economic, social, and political stability provided by World Happiness Report.
 
@@
 
-covid_2020_world <- covid_2020 %>%
+covid_2020_world <- covid_2020 |>
   filter(
@@
 # Step 1: Aggregate cyclone counts by country
-cyclone_counts <- cyclones_data %>%
-  group_by(country) %>%
+cyclone_counts <- cyclones_data |>
+  group_by(country) |>
   summarise(cyclone_count = n())
@@
 # Step 2: Join with happiness data on 'country' to add distance from the equator
-combined_data <- cyclone_counts %>%
-  inner_join(happiness, by = "country") %>%
+combined_data <- cyclone_counts |>
+  inner_join(happiness, by = "country") |>
   select(country, cyclone_count, distance_to_equator_km)
@@
 # Step 3: Remove the top 3 highest cyclone counts
-combined_data <- combined_data %>%
-  arrange(desc(cyclone_count)) %>%
+combined_data <- combined_data |>
+  arrange(desc(cyclone_count)) |>
   slice(-1:-3)
@@
 
-ggplot(cyclones_data %>% filter(!is.na(category)), aes(x = km_distance_from_equator, y = category, fill = category)) +
+ggplot(cyclones_data |> filter(!is.na(category)), aes(x = km_distance_from_equator, y = category, fill = category)) +
   geom_density_ridges(scale = 2, alpha = 0.7, color = "black", size = 0.3) +
@@
 happiness_bands_life <- happiness |>
-  mutate(distance_band = cut(distance_to_equator_km.y, breaks = seq(0, 10000, by = 500), include.lowest = TRUE, labels = seq(0, 9500, by = 500))) %>%
+  mutate(distance_band = cut(distance_to_equator_km.y, breaks = seq(0, 10000, by = 500), include.lowest = TRUE, labels = seq(0, 9500, by = 500))) |>
   group_by(distance_band) |>

### 2024-11-15 Draft.Rmd
Update Draft.Rmd

@@
 # Purpose
+
 This study aims to
@@
 ## \<\<Cyclones Data\>\>
+
 ```{r load_data1}
@@
   )
+happiness
 
@@
 
-`happiness` dataset is global census on a countries economic, social, and political stability provided by World Happiness Report.
+the `happiness` dataset was loaded, containing data on countries' economic, social, and political stability from the World Happiness Report.
+To ensure consistency across datasets, some country names were adjusted to match the naming conventions in the `world_map` dataset.
+This adjustment was necessary for accurate geographic mapping.
+After renaming, the `happiness` dataset was joined with `world_map` to incorporate latitude and longitude coordinates for each country, enabling the visualization of global data trends on a map.
+Additional geographic information was also added, including the distance of each country from the equator and hemisphere designation, by using centroids of each country's polygon in the `world` dataset.
+This enhancement facilitates spatial analysis based on proximity to the equator and hemisphere differences.
 
@@
 We explored many aspects of the data, but will demonstrate three.
-These are \<\<Geographic and environmental Factors VS a countries overall well being\>\>, \<\<Covid Cases Vs perception of corruption\>\>, and \<<Freedom to Make Life Choices & Climate Awareness (OR DISTANCE VS COVID)>\>
+These are \<\<Geographic and environmental Factors VS a countries overall well being\>\>, \<\<Covid Cases Vs perception of corruption\>\>, and \<\<Freedom to Make Life Choices & Climate Awareness (OR DISTANCE VS COVID)\>\>
 
-The analysis highlighted an interesting finding, as shown in @ref(fig
-): a significant concentration of COVID-19 cases appears within the 1500-2500 km range from the equator, suggesting a strong relationship between COVID-19 cases and cyclone activity within a margin of approximately 9 degrees of longitude.
+The analysis highlighted an interesting finding, as shown in @ref(fig ): a significant concentration of COVID-19 cases appears within the 1500-2500 km range from the equator, suggesting a strong relationship between COVID-19 cases and cyclone activity within a margin of approximately 9 degrees of longitude.
 This area also highlights a geographical strip pointing to a genreal trend of lower GDP, government stability, and overall happiness.

### 2024-11-15 Draft.Rmd
Update Draft.Rmd

@@
 
-The analysis highlighted an interesting finding, as shown in @ref(fig ): a significant concentration of COVID-19 cases appears within the 1500-2500 km range from the equator, suggesting a strong relationship between COVID-19 cases and cyclone activity within a margin of approximately 9 degrees of longitude.
+The analysis highlighted an interesting finding, as shown in \@ref(fig:insight1): a significant concentration of COVID-19 cases appears within the 1500-2500 km range from the equator, suggesting a strong relationship between COVID-19 cases and cyclone activity within a margin of approximately 9 degrees of longitude.
 This area also highlights a geographical strip pointing to a genreal trend of lower GDP, government stability, and overall happiness.
 
-```{r insight1, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
+```{r insight1, fig.cap=""}
 
@@
 
-```{r}
+```{r insight2, fig.cap=""}
 #Social support & life ladder

### 2024-11-15 Draft.Rmd
removed extra graphs added insight 2 summary stats started intro and abstract started data change desc for insight 2

--- 2024-11-14 Draft.Rmd
@@
 
-# Instructions
-
-You are encouraged to remove this instruction section prior to submission.
-
-It is recommended that you follow the structure of this template.
-The text is all placeholder - you are free to change any/all wording as you please, but it is very helpful for the grading process if you keep the same structure.
-Anything in \<<double angle brackets>\> definitely needs to be changed, but you are free to change any/all sentences!
-
-Note that all of the code is *hidden* by default. This file will be graded based on the insights, not the code.
-
-You will only submit the PDF version of this document.
-To knit to PDF, you'll need to run `install.packages("tinytex")` in the console, followed by `tinytex::install_tinytex()` (DO NOT PUT THESE COMMANDS IN AN RMD FILE!!!).
-If you encounter errors in "Knit to PDF", you can "knit to html" and then print the html file to PDF using your operating system's PDF view (e.g.
-Adobe Acrobat).
-Only standalone PDF files will be accepted by MyLS.
-
 # Purpose
-This study aims to
+This study aims to investigate the relationship between freedom of citizens (i.e.
+freedom of choices, a just government) and quality of life for citizens.
+The more freedom for citizens allows for better quality of life.
 
 # Abstract
+
+This paper will aim to show as freedom of citizens increase and citizen's quality of life improves.
+The relationship we look for is a positive correlation.
+First, we explore the relationship of social support and life expectancy.
+We observe more freedom allows for better opportunities for social support.
 
@@
 
+Quality of life determines happiness which is impacted by many factors.
+
 # Data Description
@@
 ## \<\<Cyclones Data\>\>
-```{r load_data1}
+
+```{r cyclones_data}
 cyclones_data <- read_parquet("cyclones_data.parquet")
@@
 
+For our second insight we needed to join the COVID_2020 data along with the happiness data, to accomplish this we applied the "inner_join()" function with the argument "by = continent'". This allo
+
 # Exploratory Data Analysis
@@
 
-A report conducted by the [National Librarbty of Medicine](https://pmc.ncbi.nlm.nih.gov/articles/PMC8079387/) highlights a strong correlation between COVID-19 cases and proximity to the equator, suggesting that certain geographic and environmental factors may influence virus spread and outbreak.
+A report conducted by the [National Library of Medicine](https://pmc.ncbi.nlm.nih.gov/articles/PMC8079387/) highlights a strong correlation between COVID-19 cases and proximity to the equator, suggesting that certain geographic and environmental factors may influence virus spread and outbreak.
 This observation signifies a trend where regions near the equator are subject to unique vulnerabilities, not only experiencing increased COVID-19 cases but also facing intensified cyclone activity (as shown in the figure above), volcanic activity, elevated ocean currents, seasonal flooding, and the presence of natural carbon sinks.
@@
 
-```{r}
+The next insight that we found is shown in \@ref(fig:insight2).
+
+```{r insight2, fig.height=4, fig.width=6, fig.cap="Boxplot of perception of corruption vs total COVID cases by continent"}
+
+covid_happiness <- inner_join(covid_2020_world, happiness, by = "country")
+
+filter_covid_happiness <- covid_happiness |> filter(!is.na(total_cases), !is.na(perceptions_of_corruption))
+
+ratio_covid_happiness <- filter_covid_happiness |> group_by(country) |>
+  mutate(
+    ratio = total_cases * perceptions_of_corruption
+  )
+
+covid_happiness |> filter(!is.na(total_cases), !is.na(perceptions_of_corruption), total_cases <= 20000000) |>
+  ggplot(mapping = aes(x = perceptions_of_corruption, y = total_cases, color = continent)) + geom_point() + geom_smooth(method = "lm", se = FALSE) + facet_wrap(~continent) + labs(title = "Perception of Corruption and COVID cases by continent", subtitle = "Weight of COVID cases respective to perception of corruption", x = "Perception of Corruption (0 - 1)", y = "Total Cases", caption = "Source: covid.ourworldindata.org\nworldhappiness.report") + theme_minimal() + theme(legend.position = "None")
+
+```
+
+This insight is supported by the summary statistics in table \@ref(tab:summary_stats)
+
+```{r summary_stats}
+library(gtsummary)
+library(gapminder)
+
+insight_2_sum_stats <- covid_happiness |> group_by(continent) |> select(continent, perceptions_of_corruption, total_cases) |> 
+  tbl_summary(by = continent)
+insight_2_sum_stats
+```
+
+Finally, \@ref(fig:insight3) shows ...
+
+```{r insight3, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
+happiness <- happiness|>
+  left_join(
+    rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") |>
+      transmute(
+        country = name_long,
+        distance_to_equator_km = geosphere::distGeo(
+          cbind(st_coordinates(st_centroid(geometry))[,1], st_coordinates(st_centroid(geometry))[,2]),
+          cbind(st_coordinates(st_centroid(geometry))[,1], 0)
+        ) / 1000  # Convert to km
+      ),
+    by = "country"
+  )
+
+happiness_bands_gdp <- happiness |>
+  mutate(distance_band = cut(distance_to_equator_km.y, breaks = seq(0, 10000, by = 500), include.lowest = TRUE, labels = seq(0, 9500, by = 500))) |>
+  group_by(distance_band) |>
+  summarize(avg_log_gdp_per_capita = mean(log_gdp_per_capita, na.rm = TRUE))
+
+# Plot for log GDP per capita
+ggplot(happiness_bands_gdp, aes(x = distance_band, y = avg_log_gdp_per_capita)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  annotate("rect", xmin = "1500", xmax = "2500", ymin = 0, ymax = Inf, fill = "orange", alpha = 0.8) +
+  labs(
+    title = "Average GDP per Capita by Distance from Equator",
+    x = "Distance from Equator (km, Binned)",
+    y = "Average Log GDP per Capita"
+  ) +
+  scale_x_discrete(breaks = seq(0, 9500, by = 1000)) +
+  coord_cartesian(ylim = c(min(happiness_bands_gdp$avg_log_gdp_per_capita, na.rm = TRUE) - 0.1, 
+                           max(happiness_bands_gdp$avg_log_gdp_per_capita, na.rm = TRUE) + 0.1)) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
+
+# Group data into 500 km distance bands and calculate average life ladder score for each band
+happiness_bands_life <- happiness |>
+  mutate(distance_band = cut(distance_to_equator_km.y, breaks = seq(0, 10000, by = 500), include.lowest = TRUE, labels = seq(0, 9500, by = 500))) |>
+  group_by(distance_band) |>
+  summarize(avg_life_ladder = mean(life_ladder, na.rm = TRUE))
+
+# Plot for life ladder
+ggplot(happiness_bands_life, aes(x = distance_band, y = avg_life_ladder)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  annotate("rect", xmin = "1500", xmax = "2500", ymin = 0, ymax = Inf, fill = "orange", alpha = 0.8) +
+  labs(
+    title = "Average Life Ladder Score by Distance from Equator",
+    x = "Distance from Equator (km, Binned)",
+    y = "Average Life Ladder Score"
+  ) +
+  scale_x_discrete(breaks = seq(0, 9500, by = 1000)) +
+  coord_cartesian(ylim = c(min(happiness_bands_life$avg_life_ladder, na.rm = TRUE) - 0.1, 
+                           max(happiness_bands_life$avg_life_ladder, na.rm = TRUE) + 0.1)) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
+```
+
+```{r insight2, fig.cap=""}
 #Social support & life ladder
@@
 
-This insight is supported by the summary statistics in table \@ref(tab:summary_stats)
-
-The next insight that we found is shown in \@ref(fig:insight2).
-
-```{r insight3, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
-
-covid_happiness <- inner_join(covid_2020_world, happiness, by = "country")
-
-filter_covid_happiness <- covid_happiness |> filter(!is.na(total_cases), !is.na(perceptions_of_corruption))
-
-ratio_covid_happiness <- filter_covid_happiness |> group_by(country) |>
-  mutate(
-    ratio = total_cases * perceptions_of_corruption
-  )
-
-covid_happiness |> filter(!is.na(total_cases), !is.na(perceptions_of_corruption)) |>
-  ggplot(mapping = aes(x = perceptions_of_corruption, y = total_cases, color = continent)) + geom_boxplot() + facet_wrap(~continent) + labs(title = "Perception of Corruption and COVID cases by continent", subtitle = "Weight of COVID cases repective to perception of corruption", x = "Perception of Corruption (0 - 1)", y = "Total Cases", caption = "Source: covid.ourworldindata.org\nworldhappiness.report") + theme_minimal() + theme(legend.position = "None")
-
-```
-
-Finally, \@ref(fig:insight3) shows ...
-
-```{r insight3, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
-
-```
-
 # Conclusion and Future Work

### 2024-11-15 Draft.Rmd
deleted clutter

@@
 
-The analysis highlighted an interesting finding, as shown in @ref(fig
-): a significant concentration of COVID-19 cases appears within the 1500-2500 km range from the equator, suggesting a strong relationship between COVID-19 cases and cyclone activity within a margin of approximately 9 degrees of longitude.
+The analysis highlighted an interesting finding, as shown in \@ref(fig:insight1): a significant concentration of COVID-19 cases appears within the 1500-2500 km range from the equator, suggesting a strong relationship between COVID-19 cases and cyclone activity within a margin of approximately 9 degrees of longitude.
 This area also highlights a geographical strip pointing to a genreal trend of lower GDP, government stability, and overall happiness.
 
-```{r insight1, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
+```{r insight1, fig.cap=""}
 
@@
 
-```{r insight2, fig.height=4, fig.width=6, fig.cap="Boxplot of perception of corruption vs total COVID cases by continent"}
+```{r insight2, fig.height=4, fig.width=6, fig.cap=""}
 
@@
   )
-
-happiness_bands_gdp <- happiness |>
-  mutate(distance_band = cut(distance_to_equator_km.y, breaks = seq(0, 10000, by = 500), include.lowest = TRUE, labels = seq(0, 9500, by = 500))) |>
-  group_by(distance_band) |>
-  summarize(avg_log_gdp_per_capita = mean(log_gdp_per_capita, na.rm = TRUE))
-
-# Plot for log GDP per capita
-ggplot(happiness_bands_gdp, aes(x = distance_band, y = avg_log_gdp_per_capita)) +
-  geom_bar(stat = "identity", fill = "steelblue") +
-  annotate("rect", xmin = "1500", xmax = "2500", ymin = 0, ymax = Inf, fill = "orange", alpha = 0.8) +
-  labs(
-    title = "Average GDP per Capita by Distance from Equator",
-    x = "Distance from Equator (km, Binned)",
-    y = "Average Log GDP per Capita"
-  ) +
-  scale_x_discrete(breaks = seq(0, 9500, by = 1000)) +
-  coord_cartesian(ylim = c(min(happiness_bands_gdp$avg_log_gdp_per_capita, na.rm = TRUE) - 0.1, 
-                           max(happiness_bands_gdp$avg_log_gdp_per_capita, na.rm = TRUE) + 0.1)) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
-
-# Group data into 500 km distance bands and calculate average life ladder score for each band
-happiness_bands_life <- happiness |>
-  mutate(distance_band = cut(distance_to_equator_km.y, breaks = seq(0, 10000, by = 500), include.lowest = TRUE, labels = seq(0, 9500, by = 500))) |>
-  group_by(distance_band) |>
-  summarize(avg_life_ladder = mean(life_ladder, na.rm = TRUE))
-
-# Plot for life ladder
-ggplot(happiness_bands_life, aes(x = distance_band, y = avg_life_ladder)) +
-  geom_bar(stat = "identity", fill = "steelblue") +
-  annotate("rect", xmin = "1500", xmax = "2500", ymin = 0, ymax = Inf, fill = "orange", alpha = 0.8) +
-  labs(
-    title = "Average Life Ladder Score by Distance from Equator",
-    x = "Distance from Equator (km, Binned)",
-    y = "Average Life Ladder Score"
-  ) +
-  scale_x_discrete(breaks = seq(0, 9500, by = 1000)) +
-  coord_cartesian(ylim = c(min(happiness_bands_life$avg_life_ladder, na.rm = TRUE) - 0.1, 
-                           max(happiness_bands_life$avg_life_ladder, na.rm = TRUE) + 0.1)) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
-```
-
-```{r insight2, fig.cap=""}
-#Social support & life ladder
-
-data_combined <- inner_join(covid_2020, happiness, by = "country")
-
-median_social_support <- median(data_combined$social_support, na.rm = TRUE)
-
-ggplot(data_combined, aes(x = social_support, y = life_ladder, color = continent)) +
-  geom_point(alpha = 0.7, position = "jitter") + 
-  geom_smooth(method = "lm", color = "blue", se = TRUE) + 
-  geom_vline(xintercept = median_social_support, color = "red", linetype = "dashed", linewidth = 1) + 
-  labs(
-    title = "Comparison of Social Support and Life Ladder by Continent",
-    x = "Social Support",
-    y = "Life Ladder (Happiness Score)",
-    color = "Continent"
-  ) +
-  facet_wrap(~ continent) + 
-  theme_minimal() +
-  theme(panel.grid.major = element_line(color = "grey", linetype = "dashed", linewidth = 0.5))
-
-# Social Support & Life Expectancy
-
-ggplot(data_combined, aes(x = social_support, y = healthy_life_expectancy_at_birth, color = continent)) +
-  geom_point(alpha = 0.7, position = "jitter") + 
-  geom_smooth(method = "lm", color = "blue", se = TRUE) + 
-  geom_vline(xintercept = median_social_support, color = "red", linetype = "dashed", linewidth = 1) + 
-  labs(
-    title = "Comparison of Social Support and Life Expectancy by Continent",
-    x = "Social Support",
-    y = "Healthy Life Expectancy at Birth",
-    color = "Continent"
-  ) +
-  facet_wrap(~ continent) + 
-  theme_minimal() +
-  theme(panel.grid.major = element_line(color = "grey", linetype = "dashed", linewidth = 0.5))
-
-data_combined <- inner_join(covid_2020, happiness, by = "country")
-
-median_total_cases <- median(data_combined$total_cases, na.rm = TRUE)
-
-ggplot(data_combined, aes(x = total_cases / 100, y = freedom_to_make_life_choices)) +
-  geom_point(alpha = 0.5, position = "jitter") + 
-  geom_smooth(method = "lm", color = "blue", se = TRUE) + 
-  geom_vline(xintercept = median_total_cases, color = "red", linetype = "dashed", linewidth = 1) + 
-  labs(
-    title = "Comparison of COVID-19 Cases and Freedom",
-    x = "Total COVID-19 Cases",
-    y = "Freedom"
-  ) +
-  scale_x_log10() +  
-  scale_y_log10() + 
-  theme_minimal() +
-  theme(panel.grid.major = element_line(color = "grey", linetype = "dashed", linewidth = 0.5))
-
-#Life expectancy & Cyclone count
-
-cyclone_counts <- cyclones_data |>
-  group_by(country) |>
-  summarise(cyclone_count = n())
-
-combined_data <- cyclone_counts |>
-  inner_join(happiness, by = "country") |>
-  select(country, cyclone_count, healthy_life_expectancy_at_birth)
-
-combined_data <- combined_data |>
-  arrange(desc(cyclone_count)) |>
-  slice(-1:-3)
-
-ggplot(combined_data, aes(y = cyclone_count, x = healthy_life_expectancy_at_birth)) +
-  geom_point(alpha = 0.7) +
-  geom_smooth(method = "lm", se = TRUE, color = "blue") +
-  labs(
-    title = "Relationship Between Cyclone Count and Life Expectancy (Outlier Removed)",
-    x = "Cyclone Count",
-    y = "Healthy Life Expectancy at Birth"
-  ) +
-  theme_minimal()
-
-colnames(happiness)
-
-# distance & cyclone count
-
-# Load required libraries
-library(tidyverse)
-
-# Step 1: Aggregate cyclone counts by country
-cyclone_counts <- cyclones_data |>
-  group_by(country) |>
-  summarise(cyclone_count = n())
-
-# Step 2: Join with happiness data on 'country' to add distance from the equator
-combined_data <- cyclone_counts |>
-  inner_join(happiness, by = "country") |>
-  select(country, cyclone_count, distance_to_equator_km)
-
-# Step 3: Remove the top 3 highest cyclone counts
-combined_data <- combined_data |>
-  arrange(desc(cyclone_count)) |>
-  slice(-1:-3)
-
-# Step 4: Plot the relationship between distance to equator and cyclone count
-ggplot(combined_data, aes(x = distance_to_equator_km, y = cyclone_count)) +
-  geom_point(alpha = 0.7, color = "blue") +
-  geom_smooth(method = "lm", color = "red", se = TRUE) +
-  labs(
-    title = "Comparison of Cyclone Count and Distance from Equator (Top 3 Outliers Removed)",
-    x = "Distance from Equator (km)",
-    y = "Cyclone Count"
-  ) +
-  theme_minimal()
-
-ggplot(cyclones_data |> filter(!is.na(category)), aes(x = km_distance_from_equator, y = category, fill = category)) +
-  geom_density_ridges(scale = 2, alpha = 0.7, color = "black", size = 0.3) +
-  labs(
-    x = "Distance from Equator (km)",
-    y = "Cyclone Category",
-    title = "Ridgeline Plot of Distance from Equator by Cyclone Category"
-  ) +
-  coord_cartesian(xlim = c(0, 6500)) +
-  theme_minimal() +
-  theme(
-    legend.position = "none",
-    panel.grid.major = element_line(color = "grey80"),
-    panel.grid.minor = element_blank(),
-    axis.text.y = element_text(size = 10),
-    axis.text.x = element_text(size = 10)
-  )
-
-colnames(happiness)
-
-ggplot(cyclones_data |> filter(!is.na(category)), aes(x = km_distance_from_equator, y = category, fill = category)) +
-  geom_density_ridges(scale = 2, alpha = 0.7, color = "black") +
-  annotate("rect", xmin = 1500, xmax = 2500, ymin = 0, ymax = Inf, fill = "orange", alpha = 0.8) +
-  labs(
-    x = "Distance from Equator (km)",
-    y = "Cyclone Category",
-    title = "Ridgeline Plot of Distance from Equator by Cyclone Category"
-  ) +
-  coord_cartesian(xlim = c(0, 6500)) +
-  theme_minimal() +
-  theme(
-    legend.position = "none",
-    panel.grid.major = element_line(color = "grey80"),
-    panel.grid.minor = element_blank(),
-    axis.text.y = element_text(size = 10),
-    axis.text.x = element_text(size = 10)
-  )
-
-happiness <- happiness|>
-  left_join(
-    rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") |>
-      transmute(
-        country = name_long,
-        distance_to_equator_km = geosphere::distGeo(
-          cbind(st_coordinates(st_centroid(geometry))[,1], st_coordinates(st_centroid(geometry))[,2]),
-          cbind(st_coordinates(st_centroid(geometry))[,1], 0)
-        ) / 1000  # Convert to km
-      ),
-    by = "country"
-  )
-
-happiness
 

### 2024-11-15 Draft.Rmd
Insight 3 writing

@@
 
-Finally, \@ref(fig:insight3) shows ...
+Finally, \@ref(fig:insight3) shows the impacts of countries distances from the equator on their average GDP per capita.
 
+Firstly, countries are grouped by their distance from the equator into 500km intervals to effectively display all the info. The plot then compares this info to the average GDP of each group.
+
+We can see the positive correlation between the two variables.
+As the distance from the equator begins to increase, the average GDP per capita also increases.
+There are some outliers such as the countries right at the equator and the ones at 2500km away, which is highlight by the golden bar.
+
+A big reason as to why such a correlation exists is the effect of warmer weather on the climate.
+Warmer weather makes it harder to grow crops, limits water availability, increases number of parasites and potential diseases, and overall leads to poorer health.
 ```{r insight3, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}

### 2024-11-15 Draft.Rmd
Update Draft.Rmd

@@
 # Introduction
-
-Climate change is something that has been studied. Here's some relevant information about the context of our study.
-
-If needed, this paragraph is more information about the context.
-
-In this report, we are going to explore some aspects climate change and the impact and/or perceptions of it by using exploratory techniques.
-We'll explore \<<general description of data>\> using \<<general description of techniques>\>.
-
-By the end of this report, we will have shown ...
 
@@
 
-Overall, we found \<<general ideas>\>.
+Overall, we found that geographic and environmental factors, freedom to make life choices, and perceptions of corruption play significant roles in shaping a country's stability and overall well-being.
+In particular, regions closer to the equator show greater vulnerability to crises, such as COVID-19 outbreaks and natural disasters, which negatively impact GDP and social support.
+At the same time, greater personal freedom and transparency in governance correlate with higher life satisfaction, highlighting the importance of governance in fostering well-being.
 
-A second paragraph about our findings.
+Our findings reveal a complex relationship between proximity to the equator and national challenges, as countries within certain latitude ranges face compounded vulnerabilities.
+Additionally, freedom to make life choices appears closely tied to social support and satisfaction, suggesting that individual autonomy may be essential for improving quality of life.
+The significant correlation between perceived corruption and COVID-19 cases further underscores the importance of transparent governance, especially during public health crises.
 
-The next steps in this analysis are...
+The following steps in this analysis expand our approach by exploring regional variations within countries, considering the effects of policy measures, and assessing additional factors like urbanization and healthcare access.
+These expansions could provide a more nuanced understanding of the diverse factors affecting national well-being.
+Further research into causal links between geographic location and economic or social outcomes could deepen insights into how environmental and governance factors interact.
 
-The limitations of this analysis are as follows.
-(Do not simply list potential issues with sampling, but relate them to your analysis and how they affect your conclusions.
-An honest and complete acknowledgement of the limitations makes the analysis more trustworthy.)
+The limitations of this analysis include our reliance on national-level data, which may obscure subnational disparities that could offer a more detailed picture.
+Additionally, our analysis captures a specific moment, missing how policy shifts or environmental changes could alter these trends.
+Cultural differences, potential measurement biases, and the correlational nature of our findings limit the ability to establish causal links, as unmeasured factors may play significant roles in our observed trends.
+Recognizing these limitations makes our conclusions more robust and emphasizes the complex dynamics of well-being across global contexts.
 

### 2024-11-15 Draft.Rmd
Update Draft.Rmd

@@
 
-`happiness` dataset is global census on a countries economic, social, and political stability provided by World Happiness Report.
+The `happiness` dataset was loaded, containing data on countries, economic, social, and political stability from the World Happiness Report.
+Some country names were adjusted to match the naming conventions in the `world_map` dataset to ensure consistency across datasets.
+This adjustment was necessary for accurate geographic mapping.
+
+After renaming, the `happiness` dataset was joined with `world_map` to incorporate latitude and longitude coordinates for each country, enabling the visualization of global data trends on a map.
+Additional geographic information, including the distance of each country from the equator and hemisphere designation, was also added by using the centroids (geographical center) of each country's polygon in the `world` dataset.
+This enhancement facilitates spatial analysis based on proximity to the equator and hemisphere differences.
 

### 2024-11-15 Draft.Rmd
brushed

--- 2024-11-15 Draft.Rmd
@@
 The relationship we look for is a positive correlation.
-First, we explore the relationship of social support and life expectancy.
-We observe more freedom allows for better opportunities for social support.
+First, for the first insight, we explore the relationship of cyclone frequencies and geological location.
+Cyclones of higher category appear more frequently near the equator (~2000 km from the equator), more freedom allows for an individual to choose where they can locate themselves and those who choose to live farther away from the equator may feel less burdened by constant cyclones.
+Second, a more just government allows for a citizen to live with more freedom, from th second insight, "perception of corruption" for a government and "COVID cases" shows us for some continents, there are lower cases of COVID when the government is perceived to be fair.
+Though there also exists a negative correlation, where continents (i.e.
+Asia) with more corrupt ruling powers have lower total count of COVID cases.
+Third, again back to the notion of choosing where to live, from the third insight, we observe average GDP in relation to distance from the equator.
+Areas farther from the equator tend to have higher on average higher GDP than those closer to the equator.
+We now show evidence to further support our original purpose.
+That freedom of citizens in fact impacts the quality of life for citizens.
 
@@
 
-Climate change is something that has been studied. Here's some relevant information about the context of our study.
+Quality of life determines happiness which is impacted by many factors.
+Some important factors are environmental and socioeconomic factors.
+Environmentally we must keep in mind location, since different countries in different continents vary in climate, geographic identity, etc.
+Whilst socioeconomically, different countries have different forms of government, culture.
 
-If needed, this paragraph is more information about the context.
+In this report, we will explore the geographical aspect and governance aspect of freedom and what do these factors impact on an individual's daily life through exploring and digesting our data.
+We will explore data; COVID cases during the year 2020, cyclones from NOAA (National Oceanic and Atmospheric Administration), happiness data from (World Happiness Report).
+We apply uni-variate and bi-variate analysis, specifically for the bard chart and ridge line graph we care about the frequency (weight) of our independent variables, for the scatter plot to determine the relationship between our variable x and y is applying bivariate analysis.
 
-In this report, we are going to explore some aspects climate change and the impact and/or perceptions of it by using exploratory techniques.
-We'll explore \<<general description of data>\> using \<<general description of techniques>\>.
-
-By the end of this report, we will have shown ...
-
-Quality of life determines happiness which is impacted by many factors.
+By the end of the report, we will have shown in general that as freedom increases for citizens and individuals, so to does quality of life.
 
@@
 
-`happiness` dataset is global census on a countries economic, social, and political stability provided by World Happiness Report.
+The `happiness` data set is global census on a countries economic, social, and political stability provided by World Happiness Report.
+From part 1, extra steps were applied to clean the data set.
+Some steps taken are renaming countries to allow for easy joining of data sets (happiness dataset and world map).
+For this case only specific columns were needed and hence selected specific columns.
 
@@
 
-Explain how any combinations of data were performed. Explain what kind of join was needed, whether columns had to be modified (for example, matching "country" names.)
-
-For our second insight we needed to join the COVID_2020 data along with the happiness data, to accomplish this we applied the "inner_join()" function with the argument "by = continent'". This allo
+For the first data set, Cyclones data "sf_join" was used to create geographical data for a map for distances of cyclones from the equator (this was not used but is in the appendix).
+Only necessary columns were selected from the data set.
+For the second data set, World data was used from 'rnaturalearth' package.
+First countries were renamed to allow for easier joining of data sets.
+New columns were created for the world map to be created.
+Then the data set is combined with COVID_2020 by country.
+For the third data set, country names were changed for easier merge.
+Then applying "right_join" to keep the data from world_map and combine with happiness report data.
+This again is combined with world data, this time applying "left_join" to keep the data from happiness data and combine with world data.
 
@@
 We explored many aspects of the data, but will demonstrate three.
-These are \<\<Geographic and environmental Factors VS a countries overall well being\>\>, \<\<Covid Cases Vs perception of corruption\>\>, and \<<Freedom to Make Life Choices & Climate Awareness (OR DISTANCE VS COVID)>\>
+These are \<\<Geographic and environmental Factors VS a countries overall well being\>\>, \<\<Covid Cases Vs perception of corruption\>\>, and \<<Life ladder and GDP vs Distance from Equator>\>
 
 The analysis highlighted an interesting finding, as shown in \@ref(fig:insight1): a significant concentration of COVID-19 cases appears within the 1500-2500 km range from the equator, suggesting a strong relationship between COVID-19 cases and cyclone activity within a margin of approximately 9 degrees of longitude.
-This area also highlights a geographical strip pointing to a genreal trend of lower GDP, government stability, and overall happiness.
+This area also highlights a geographical strip pointing to a general trend of lower GDP, government stability, and overall happiness.
 
 ```{r insight1, fig.cap=""}
-
-covid_2020_world <- covid_2020 |>
-  filter(
-    total_cases < 50000000 
-  )
-
-ggplot(covid_2020_world, aes(x = distance_to_equator_km, y = total_cases)) +
-  geom_point(color = "steelblue", alpha = 0.3) +  # Lower alpha for more transparency
-  geom_density_2d(color = "darkorange") +  # Density lines to show clustering
-  geom_vline(xintercept = 1500, linetype = "dashed", color = "black") +
-  geom_vline(xintercept = 2500, linetype = "dashed", color = "black")  +
-  labs(
-    title = "Scatter Plot of COVID-19 Cases by Distance from Equator",
-    x = "Distance from Equator (km)",
-    y = "Total COVID-19 Cases"
-  ) +
-  coord_cartesian(xlim = c(0, 10250)) +
-  scale_y_continuous(
-    labels = scales::comma,
-    trans = "log10",
-    breaks = c(1, 10, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7, 5e7)  
-  ) +
-  theme_minimal()
 
@@
 
-covid_happiness <- inner_join(covid_2020_world, happiness, by = "country")
+covid_happiness <- inner_join(covid_2020, happiness, by = "country")
 
@@
 
-```{r summary_stats}
+```{r summary_stats, fig.height=4, fig.width=6,}
 library(gtsummary)
@@
 
-Finally, \@ref(fig:insight3) shows the impacts of countries distances from the equator on their average GDP per capita.
+From [Scientific Reports (Sci Rep)](https://www.nature.com/articles/s41598-021-02802-1), the paper implies that corruption leads to lower rates of vaccination.
+This can result in higher counts of COVID cases.
+Also [Globalization and Health](https://globalizationandhealth.biomedcentral.com/articles/10.1186/s12992-023-00994-x) also accounts to how corruption can manifest due to catastrophic events such as a pandemic.
+Then corruption affects how we tackle on health issues and health issues also affect how corruption is procured in the globlal stage.
 
-Firstly, countries are grouped by their distance from the equator into 500km intervals to effectively display all the info. The plot then compares this info to the average GDP of each group.
+Finally, \@ref(fig:insight3) shows ...
 
-We can see the positive correlation between the two variables.
-As the distance from the equator begins to increase, the average GDP per capita also increases.
-There are some outliers such as the countries right at the equator and the ones at 2500km away, which is highlight by the golden bar.
-
-A big reason as to why such a correlation exists is the effect of warmer weather on the climate.
-Warmer weather makes it harder to grow crops, limits water availability, increases number of parasites and potential diseases, and overall leads to poorer health.
-```{r insight3, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
+```{r insight3, fig.height=4, fig.width=6, fig.cap=""}
 happiness <- happiness|>
@@
 
-# Group data into 500 km distance bands and calculate average life ladder score for each band
-happiness_bands_life <- happiness |>
-  mutate(distance_band = cut(distance_to_equator_km.y, breaks = seq(0, 10000, by = 500), include.lowest = TRUE, labels = seq(0, 9500, by = 500))) |>
-  group_by(distance_band) |>
-  summarize(avg_life_ladder = mean(life_ladder, na.rm = TRUE))
-
-# Plot for life ladder
-ggplot(happiness_bands_life, aes(x = distance_band, y = avg_life_ladder)) +
-  geom_bar(stat = "identity", fill = "steelblue") +
-  annotate("rect", xmin = "1500", xmax = "2500", ymin = 0, ymax = Inf, fill = "orange", alpha = 0.8) +
-  labs(
-    title = "Average Life Ladder Score by Distance from Equator",
-    x = "Distance from Equator (km, Binned)",
-    y = "Average Life Ladder Score"
-  ) +
-  scale_x_discrete(breaks = seq(0, 9500, by = 1000)) +
-  coord_cartesian(ylim = c(min(happiness_bands_life$avg_life_ladder, na.rm = TRUE) - 0.1, 
-                           max(happiness_bands_life$avg_life_ladder, na.rm = TRUE) + 0.1)) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
 ```

### 2024-11-15 Draft.Rmd

@@
 
-```{r insight1, fig.cap=""}
+```{r insight1, fig.cap="Cyclone catergory frequency by distance from equator"}
 
@@
 
-```{r insight2, fig.height=4, fig.width=6, fig.cap=""}
+```{r insight2, fig.height=4, fig.width=6, fig.cap="Covid cases by perception of corruption"}
 
@@
 
-```{r insight3, fig.height=4, fig.width=6, fig.cap=""}
+```{r insight3, fig.height=4, fig.width=6, fig.cap="Average GDP by distance from equator"}
 happiness <- happiness|>
@@
 
+Firstly, countries are grouped by their distance from the equator into 500km intervals to effectively display all the info. The plot then compares this info to the average GDP of each group.
+
+We can see the positive correlation between the two variables.
+As the distance from the equator begins to increase, the average GDP per capita also increases.
+There are some outliers such as the countries right at the equator and the ones at 2500km away, which is highlight by the golden bar.
+
+A big reason as to why such a correlation exists is the effect of warmer weather on the climate.
+Warmer weather makes it harder to grow crops, limits water availability, increases number of parasites and potential diseases, and overall leads to poorer health.
+
 # Conclusion and Future Work
 
-Overall, we found \<<general ideas>\>.
+Overall, we found that geographic and environmental factors, freedom to make life choices, and perceptions of corruption play significant roles in shaping a country's stability and overall well-being.
+In particular, regions closer to the equator show greater vulnerability to crises, such as COVID-19 outbreaks and natural disasters, which negatively impact GDP and social support.
+At the same time, greater personal freedom and transparency in governance correlate with higher life satisfaction, highlighting the importance of governance in fostering well-being.
 
-A second paragraph about our findings.
+Our findings reveal a complex relationship between proximity to the equator and national challenges, as countries within certain latitude ranges face compounded vulnerabilities.
+Additionally, freedom to make life choices appears closely tied to social support and satisfaction, suggesting that individual autonomy may be essential for improving quality of life.
+The significant correlation between perceived corruption and COVID-19 cases further underscores the importance of transparent governance, especially during public health crises.
 
-The next steps in this analysis are...
+The following steps in this analysis expand our approach by exploring regional variations within countries, considering the effects of policy measures, and assessing additional factors like urbanization and healthcare access.
+These expansions could provide a more nuanced understanding of the diverse factors affecting national well-being.
+Further research into causal links between geographic location and economic or social outcomes could deepen insights into how environmental and governance factors interact.
 
-The limitations of this analysis are as follows.
-(Do not simply list potential issues with sampling, but relate them to your analysis and how they affect your conclusions.
-An honest and complete acknowledgement of the limitations makes the analysis more trustworthy.)
+The limitations of this analysis include our reliance on national-level data, which may obscure subnational disparities that could offer a more detailed picture.
+Additionally, our analysis captures a specific moment, missing how policy shifts or environmental changes could alter these trends.
+Cultural differences, potential measurement biases, and the correlational nature of our findings limit the ability to establish causal links, as unmeasured factors may play significant roles in our observed trends.
+Recognizing these limitations makes our conclusions more robust and emphasizes the complex dynamics of well-being across global contexts.
 

### 2024-11-15 Draft.Rmd
moved package loading to top

@@
 library(ggridges)
+library(gtsummary)
+library(gapminder)
+library(glue)
 ```
@@
 ```{r summary_stats, fig.height=4, fig.width=6,}
-library(gtsummary)
-library(gapminder)
-
 insight_2_sum_stats <- covid_happiness |> group_by(continent) |> select(continent, perceptions_of_corruption, total_cases) |> 

### 2024-11-15 A3-Pt1.Rmd
update

@@
 -   Andy (169081356)
+-   Aditya (169027493)
 -   Dibu (169708451)
@@
         country == "Bolivia (Plurinational State of)" ~ "Bolivia",
-        country == "Congo (Democratic Republic of the)" ~ "Congo, Rep.",
+        country == "Congo (Democratic Republic of the)" ~ "Congo, Dem. Rep.",
         country == "Egypt" ~ "Egypt, Arab Rep.",
